Acidocalcisome-like vacuoles constitute a feedback-controlled phosphate buffering system for the cytosol

Version 1

Abstract

Abstract

Cells experience strong variations in the consumption and availability of inorganic phosphate (Pi). Since Pi is an essential macronutrient but excess Pi has negative impacts on nucleotide hydrolysis and metabolism, its concentration must be maintained in a suitable range. Conserved storage organelles, acidocalcisomes, provide this buffering function. We used acidocalcisome-like yeast vacuoles to study how such organelles are set up to for this task. Our combined in vitro and in vivo analyses revealed that their ATP-driven polyphosphate polymerase VTC converts cytosolic Pi into inorganic polyphosphates (polyP), which it transfers into the vacuole lumen. Luminal polyphosphatases immediately hydrolyse this polyP to establish a growing reservoir of vacuolar Pi. Product inhibition by this Pi pool silences the polyphosphatases, caps Pi accumulation, and favours vacuolar polyP storage. Upon cytosolic Pi scarcity, the declining inositol pyrophosphate levels activate the vacuolar Pi exporter Pho91 to replenish cytosolic Pi. In this way, acidocalcisome-like vacuoles constitute a feedback-regulated buffering system for cytosolic Pi, which the cells can switch between Pi accumulation, Pi release, and high-capacity phosphate storage through polyP.

Introduction

Cells actively manage the concentration of Pi in their cytosol because they must strike a balance between conflicting goals (Austin & Mayer, 2020). On the one hand, Pi is a product of nucleotide-hydrolysing reactions. Its concentration has a significant impact on the free energy that these reactions can provide to drive metabolism. Since, for his reason, excessive Pi concentrations might stall metabolism, we can expect that cytosolic Pi must remain limited. On the other hand, Pi is an essential macronutrient. As a major constituent of nucleic acids and phospholipids, and as an important modifier of proteins, carbohydrates and many metabolites, it is consumed in large quantities for anabolic reactions. This can lead to situations where progression of the S-phase of the cell cycle is limited by the Pi uptake capacity of the cells (Bru et al, 2016, 2017; Gillies et al, 1981). Rapidly dividing cells, such as yeast cells, remedy this problem by maintaining phosphate stores in the form of inorganic polyP. PolyPs are chains of Pi linked through phosphoric anhydride bonds, which can be stored in acidocalcisome-like vacuoles (Urech et al, 1978; Okorokov et al, 1980). These polyP stores are accessed by cells when Pi becomes limiting in the environment, or when they face a sudden enhanced Pi consumption, e.g. during metabolic transitions from respiratory growth to fermentation, which requires much more phosphorylated metabolites (Shirahama et al, 1996; Thomas & O’Shea, 2005; Gillies et al, 1981; Nicolay et al, 1982, 1983). In such cases, cellular polyphosphate content decreases. It has therefore been proposed that polyPs are re-converted into Pi to buffer shortages in cytosolic Pi (Nicolay et al, 1983, 1982).

While this hypothesis is straightforward, we are lacking a coherent concept of how a Pi buffer based on an acidocalcisome-like organelle might work. Pioneering work on these organelles has revealed several characteristic and conserved features, such as their acidity, and their high content of basic amino acids and divalent cations (Docampo & Huang, 2016; Docampo, 2024). Of direct relevance to Pi homeostasis is the high capacity of acidocalcisome-like organelles for storing polyP. PolyP can vary greatly in length, from two to hundreds of phosphate units. The membrane of acidocalcisome-like organelles can carry polyP polymerases, such as VTC, and Pi transporters, such as Pho91 (Jimenez & Docampo, 2015; Hürlimann et al, 2007; Wang et al, 2015; Huang & Docampo, 2015; Gerasimaite & Mayer, 2016; Müller et al, 2003, 2002). VTC is a coupled polyP polymerase and translocase (Gerasimaite et al, 2014), which synthesizes polyP by transferring the ψ-phosphate of cytosolic ATP onto an elongating polyP chain (Hothorn et al, 2009). PolyP is synthesized by a catalytic domain in the centre of the Vtc4 subunit of the complex (Hothorn et al, 2009). The activity of this subunit is controlled through the SPX domains of VTC (Wild et al, 2016), which may associate into a dimerized inactive state (Pipercevic et al, 2023). They can be released from this state through the action of inositol pyrophosphates, signalling molecules that accumulate when Pi is abundant in the cytosol and signal that Pi storage in the form of polyP should be activated. The Pi state of the cytosol is communicated to VTC through a specific inositol pyrophosphate, 1,5-IP8 (Gerasimaite et al, 2017; Chabert et al, 2023).

The postulated direct channeling of polyP from the site of its synthesis through the vacuolar membrane (Gerasimaite et al, 2014) is facilitated through the structure of the VTC complex. The catalytic domain of VTC has its exit for polyP right at the entry of a polyP-conducting channel, which is formed through the transmembrane domains of the VTC complex itself (Liu et al, 2023; Guan et al, 2023; Müller et al, 2002, 2003; Gerasimaite et al, 2014). This channel was proposed to exist in an open and closed conformation and be gated through polyP (Liu et al, 2023). PolyP translocation may be driven by the electrochemical potential across the membrane, which could move the highly negatively charged polyP chain through electrophoresis. This could explain why efficient polyP synthesis depends on proton pumps such as the V-ATPase or H+-pumping pyrophosphatases (Gerasimaite & Mayer, 2016; Gerasimaite et al, 2014; Freimoser et al, 2006; Lemercier et al, 2004).

The transporter Pho91 was proposed to export vacuolar Pi (Hürlimann et al, 2007). Several properties of yeast Pho91 and its homologs from other organisms are consistent with this view. Its ablation increases vacuolar Pi and polyP content, in yeast as well as in trypanosomes (Hürlimann et al, 2007; Jimenez & Docampo, 2015; Farofonova et al, 2023), and it weakly induces the phosphate starvation program in yeast (Pinson et al, 2004), suggesting that it may induce cytosolic Pi scarcity. Furthermore, patch-clamp analyses of Pho91-like channels from yeast, trypanosomes and plants have shown their Pi permeability and the dependence of their directionality on a pH gradient across the membrane (Potapenko et al, 2018, 2019; Wang et al, 2015). In the presence of a pH gradient, the Pho91 homolog from rice, OsSpx-MFS3, mediates Pi flux along the proton gradient, consistent with a function in exporting Pi from the acidic lumen of vacuoles (Wang et al, 2015) towards the cytosol.

Acidocalcisome-like organelles also contain polyphosphatases in their lumen (Lander et al, 2016; Gerasimaite & Mayer, 2016; Kulakovskaya et al, 2021; McCarthy & Downey, 2023), which can convert polyP back into Pi. In baker’s yeast, two vacuolar polyphosphatases are known, Ppn1 and Ppn2 (Gerasimaite & Mayer, 2017; Sethuraman et al, 2001; Andreeva et al, 2019). This means that a chain of polyP, when being synthesized by VTC and arriving in the vacuolar lumen, is immediately exposed to hydrolytic enzymes that will degrade it. While this seems at first sight paradoxical, we explored the hypothesis that the co-existence of polyP-synthesizing and polyP-hydrolysing activities might be a key feature conveying to acidocalcisome-like organelles the capacity to buffer cytosolic Pi. That these organelles have a critical role to play in this process is illustrated by observations in yeast, where artificial up- or down-regulation of vacuolar polyP synthesis suffices to drive the cytosol into a state of Pi starvation or Pi excess, respectively (Desfougères et al, 2016a). Furthermore, the presence of polyP reserves delays the activation of the transcriptional phosphate starvation response, the PHO pathway (Thomas & O’Shea, 2005). We hence explored the capacity of isolated yeast vacuoles to interconvert polyP and Pi, and we characterized the roles played by the vacuolar polyphosphatases Ppn1 and Ppn2 and the vacuolar Pi transporter Pho91. Our observations can be combined with previous findings to yield a coherent model of how an acidocalcisome-like organelle can operate as a Pi buffer for the cytosol.

Results

We explored the interplay of VTC, polyphosphatases and Pho91 in the accumulation of polyP and Pi inside vacuoles using an in vitro system with purified organelles. Vacuoles can be isolated in intact form when the cells are gently opened by enzymatic digestion of the cell wall and disruption of the cell membrane by low concentrations of DEAE-dextran (Dürr et al, 1975). Organelles isolated in this way can perform many of their normal cellular functions, such as membrane fusion, membrane fission and autophagy (Michaillat et al, 2012; Sattler & Mayer, 2000; D’Agostino & Mayer, 2019; Kunz et al, 2004). They also contain active polyP polymerase (VTC) and active polyphosphatases (Ppn1 and Ppn2) and they can synthesize and import polyP (Gerasimaite et al, 2014; Gerasimaite & Mayer, 2017; Gerasimaite et al, 2017).

Vacuolar Pi accumulation depends on polyP synthesis

Purified vacuoles were incubated with an ATP-regenerating system under conditions that allow these organelles to synthesize polyP in vitro (Gerasimaite et al, 2014). After different periods of incubation, the organelles were sedimented and solubilized in detergent. Vacuolar accumulation of polyP was assayed through DAPI fluorescence and Pi was measured through malachite green. VTC is stimulated by a variety of inositol pyrophosphates, most efficiently by 1,5-InsP8 (Gerasimaite et al, 2014, 2017). Since 1,5-InsP8 is not commercially available, we used saturating concentrations of 5-InsP7 for our experiments (Pavlovic et al, 2015). This compound stimulates VTC with a higher EC50 but to the same maximal activity as 1,5-InsP8, and it was more readily available to us for routine experiments. Under the conditions used here, the organelles rapidly and efficiently produced polyP (Fig. 1), in line with previous results (Gerasimaite et al, 2014, 2017). This production was stimulated by 5-InsP7, reaching 0.7 nmol of phosphate units per µg of vacuolar protein and min. This means that already within 10 min, wildtype vacuoles produced a mass of polyP equivalent to their total protein content, indicating how efficiently these organelles synthesize polyP. This signal was entirely dependent on VTC, because it was not observed in a mutant lacking the catalytic subunit of VTC (vtc41).

VTC- and 5-IP7-dependent polyP synthesis by isolated vacuoles

Vacuoles were isolated from logarithmic cultures of BY4742 wildtype cells (BY) or from isogenic vtc4τι cells strains. They were incubated in polyphosphate synthesis assays without (A) or (B) in presence of 50 µM 5-InsP7. At indicated times, aliquots were withdrawn, solubilized in Triton X-100 and polyphosphate was quantified through the polyP-dependent fluorescence of added DAPI. Means ± SEM (standard error of the mean) of at least three independent experiments are shown.

Isolated wildtype vacuoles also accumulated Pi with significant efficiency, at an initial rate of at least 0.3 nmol µg−1 min−1 (Fig. 2A). Vacuoles from vtc41 cells showed only 10% of this Pi signal, which was not time-dependent and hence may represent a background signal from the organelle preparation. In sum, vacuolar Pi accumulation depended on polyP synthesis through VTC.

VTC-dependent accumulation of Pi in isolated vacuoles

Vacuoles were isolated from the indicated, logarithmically growing cells. PolyP overproduction was achieved either through (A, B) overexpressing VTC5 from the strong GPD promotor, or (C,D) by expressing the hyperactivating vtc3k126A and vtc4K129A alleles form their native promotors as the sole source of these two proteins. The vacuoles were incubated with an ATP-regenerating system and 50 µM 5-InsP7 under conditions allowing polyP synthesis. At indicated times, the vacuoles were solubilized with Triton X-100 and Pi was assayed through malachite green. Graphs represent the mean ± SEM of at least three independent experiments for each strain.

The kinetics of polyP and Pi accumulation were remarkably different. Whereas polyP continued to accumulate throughout the entire incubation period, Pi accumulated rapidly in the initial phase but reached a plateau within 30 min (Fig. 2A, B). Thus, although polyP production was essential for Pi accumulation, it could not become the limiting factor for it in the second phase. This is further illustrated by Pi accumulation in strains showing enhanced polyP synthesis, such as overexpressors of VTC5 (Fig. 2A, B) or strains carrying hyperactivating substitutions in the SPX domains of VTC (vtc3K126A vtc4K129A; Fig. 2 C, D) (Desfougères et al, 2016b; Wild et al, 2016).

Vacuoles from these strains accumulated polyP at a two to seven times higher initial rate than the wildtype and to higher concentrations. Nevertheless, their accumulation of Pi arrested at the same level as that of wildtype vacuoles, at 5 nmol µg−1 vacuolar protein. To estimate the corresponding luminal concentration of Pi, we measured the diameters of 100 isolated wildtype vacuoles and counted the number of vacuoles per µg of vacuolar protein. With an average diameter of 0.8 µm and 5*107 vacuoles/µg of protein we can estimate that Pi accumulation in vacuoles incubated with 5-InsP7 reached a plateau at a luminal concentration of around 30 mM. This is a substantial concentration, which is in the range of the in vivo Pi concentration of 25 mM that was measured by 31P-NMR spectroscopy of yeast cells under conditions where vacuolar Pi dominates the signal (Okorokov et al, 1980). Since, upon Pi scarcity, cytosolic Pi drops to 1 mM (Okorokov et al, 1980), a substantial Pi concentration gradient across the vacuolar membrane could drive rapid replenishment of the cytosol from a readily accessible vacuolar pool of Pi.

Vacuolar Pi accumulation is limited by feedback inhibition of Ppn1 and Ppn2

We asked why vacuolar Pi accumulation quickly forms a plateau whereas vacuolar polyP continues to accumulate. To this end, we explored the hypothesis that vacuolar polyphosphatases become product-inhibited once 30 mM of their product, Pi, has accumulated. To test polyphosphatase activity we liberated Ppn1 and Ppn2 from isolated vacuoles by detergent lysis, followed by incubation with synthetic polyP as a substrate. When incubated in the presence of Zn2+, which permits activity of Ppn1 and Ppn2 (Gerasimaite & Mayer, 2017), the substrate was consumed in less than 3 min (Fig. 3). Degradation was delayed in vacuoles from ppn11 or ppn21 mutants, and it was suppressed in vacuoles from a ppn11 ppn21 mutant, in which both polyphosphatases were ablated. In incubations with only Mg2+ instead of Zn2+, which stimulates the activity of Ppn1 but much more than that of Ppn2, polyP degradation was slower and genetic ablation of Ppn1 sufficed to stabilize polyP. The addition of 30 mM potassium phosphate attenuated degradation of polyP 5- to 10-fold, both through Ppn1 (assayed in ppn21 and in samples without Zn2+) and through Ppn2 (assayed in ppn11). That both polyphosphatases are inhibited by Pi at this concentration is consistent with the notion that product inhibition of Ppn1 and Ppn2 might limit the conversion of polyP into Pi in the vacuolar lumen and thus define the concentration of the vacuolar Pi reservoir.

Effect of Pi on polyphosphatase activity.

Vacuoles were isolated from the indicated, logarithmically growing strains. The organelles were diluted in polyphosphatase assay reaction buffer, which contained 0.1% Triton X-100 and hence liberated the luminal polyphosphatases. This lysate was incubated with polyP300 as a substrate and with A) 1 mM MgCl2 or B) 1 mM ZnCl2. Where indicated, 30 mM K-Pi pH 6.8 had been added. After the indicated times of incubation, the remaining polyP was quantified through DAPI. The DAPI signal at the beginning of the incubation served as 100% reference. Graphs represent the means ± SEM of three independent experiments.

Pho91 limits vacuolar Pi accumulation in an inositol pyrophosphate-dependent manner

To allow vacuoles to function as a Pi buffer for the cytosol, the Pi in the vacuolar lumen should become accessible in a regulated manner. The Pi transporter Pho91 is a prime candidate for mediating regulatable efflux because an overexpressed GFP fusion was reported to localize to vacuoles and Pho91 is inactivated by InsPPs through its SPX domain (Potapenko et al, 2018, 2019; Wang et al, 2015; Hürlimann et al, 2007). Since overexpression of membrane proteins in yeast easily leads to their erroneous accumulation in vacuoles, we re-investigated Pho91 localization in the absence of a strong, overexpressing promotor. When we expressed C- or N-terminal Pho91-GFP fusions from the authentic PHO91 promotor, we observed different and complex localization patterns (see Supplementary Figure 1 for examples).

Effects of N- and C-terminal fluorescent protein tags on the localization of Pho91.

PHO91 was fused with a variety of N- or C-terminal protein tags and peptide spacers as indicated. They were expressed from the endogenous PHO91 promotor or, where indicated, from ADH1 or GPD1 promotors. Cells were logarithmically grown overnight in SC medium, harvested at OD600nm of 1-2, and analyzed by fluorescence microscopy.

These localization patterns depended on the nature of the fluorescent tag and the linker peptides used to attach it to Pho91, and they often showed significant accumulation in the ER. This suggests that the fluorescent protein tags interfere with intracellular trafficking of Pho91. We hence analysed purified vacuoles to test whether the non-tagged, endogenous Pho91 is a vacuolar protein. Mass spectrometry was used to determine the enrichment of proteins in the vacuolar fraction relative to a whole cell extract (Suppl. Table 1). Peptides from Pho91 were enriched in the purified vacuoles to a similar degree (39-fold) as peptides from vacuolar marker proteins, such as the vacuolar polyP polymerase subunit Vtc3 (35-fold), the vacuolar amino acid transporter Avt3 (48-fold), the v-ATPase subunit Vma9 (46-fold), or the alkaline phosphatase Pho8 (34-fold). By contrast, typical plasma membrane proteins were barely enriched, such as the iron permease Ftr1 (2.5-fold), the Pi importer Pho87 (2.4-fold) or the polyamine importer Tpo5 (2.2-fold) (Supplementary data file 1). This co-enrichment suggests that the major fraction of non-tagged Pho91 indeed resides in the vacuole. In line with such an intracellular localization, Pho91 is the only one of the five known yeast Pi transporters that cannot feed the cells with Pi. This was tested in a quintuple knockout strain lacking these five Pi transporters (Pho84, Pho87, Pho89, Pho90, Pho91), which was kept alive through expression of Pho89 from a URA-based plasmid. Plasmid shuffling allowed to exchange this plasmid against others expressing an individual Pi transporter. Whereas individual expression of the plasma-membrane Pi transporters Pho84, Pho87, Pho89, and Pho90 supports normal colony formation (Wykoff et al, 2007), Pho91 did not support growth, at all (Fig. S2). This suggests that Pho91 is indeed a vacuolar Pi transporter but that C- or N-terminal protein fusions interfere with correct sorting and cannot serve as reliable reporters for localization of this protein.

Pho91 cannot replace other Pi transporters to support growth of yeast.

We generated a BY4741 strain with a quintuple deletion of the known Pi transporters Pho84, Pho87, Pho89, Pho90 and Pho91 (D5m). These cells were kept alive by expressing the plasma membrane Pi transporter Pho89 from a URA3-based centromeric (single copy) plasmid (pRS416). Pho91 was expressed from a HIS3-based centromeric plasmid (pRS315). Cells were plated in a dilution series on SC lacking histidine (SC-HIS) to verify that the cells had the HIS3-based PHO91 plasmid, or on SC with 5-fluoro-orotic acid (5-FOA), a drug that forces cells to lose the URA3-based pRS416 and thus to live without the Pho89 transporter. Pho91 as the sole Pi transporter (on SC + 5-FOA) does not allow cells to grow.

We tested the impact of Pho91 on Pi accumulation by vacuoles in vitro, using the same approach as above (Fig. 4). Vacuoles lacking Pho91 (pho911) accumulated Pi two times faster than the wildtype and the maximal accumulated concentration was two times higher. This is consistent with a function of Pho91 as vacuolar Pi exporter (Hürlimann et al, 2007; Potapenko et al, 2018; Wang et al, 2015).

Accumulation of Pi in isolated vacuoles

VTC- and Pho91-dependence. Vacuoles were isolated from the indicated, logarithmically growing strains. The purified organelles were incubated as in Fig. 1, i.e. in a buffer with an ATP-regenerating system that allows the synthesis of polyP, and either without (w/o) or in the presence of 50 µM 5-IP7. After the indicated periods of incubation at 27°C, an 80 µl aliquot was withdrawn, the vacuoles were sedimented by centrifugation, washed and then lysed. Released vacuolar Pi was determined by malachite green assay. Graphs represent the mean ± SEM of at least three independent experiments for each strain.

The difference between pho911 and wildtype vacuoles vanished when the vacuoles were incubated in the presence of the inositol pyrophosphate 5-InsP7. Since, as we showed above, vacuolar Pi accumulation depends on polyP synthesis through VTC, the relative enhancement of Pi accumulation in wildtype vacuoles through 5-InsP7 could be caused by downregulation of Pho91, which would make the wildtype vacuoles behave similarly as pho911 vacuoles. Alternatively, 5-InsP7 might stimulate polyP synthesis in wildtype more than in pho911 vacuoles. We could rule out the latter explanation based on two observations: An assay of polyP synthesis during the incubation (Fig. 5) revealed that 5-InsP7 stimulated polyP synthesis in wildtype and pho911 vacuoles to similar degrees. Furthermore, as shown above (Fig. 2), polyP synthesis activity in wildtype cells is not rate-limiting for vacuolar Pi accumulation. Therefore, we attribute the enhancement of Pi accumulation through 5-InsP7 to Pho91. Upon Pi scarcity, the declining inositol pyrophosphate levels should then activate Pho91 to replenish cytosolic Pi.

PolyP accumulation in pho91 mutant vacuoles.

Vacuoles from wildtype (BY) and isogenic pho91τι cells were isolated and incubated under conditions supporting polyP synthesis and Pi accumulation as in Fig. 4, in the absence (w/o) or presence of 50 µM 5-IP7. At the indicated timepoints, aliquots were withdrawn, the vacuoles were lysed in detergent, and polyP was assayed through DAPI fluorescence.

Interference with vacuolar polyP turnover provokes cytosolic Pi scarcity

We tested the effects of this postulated cycle of polyP synthesis, polyphosphatase activity and Pho91-mediated Pi export on Pi homeostasis in vivo. Pho4-GFP was used as a reporter, because this transcription factor shuttles between nucleus and cytosol. Under cellular Pi scarcity and correspondingly low inositol pyrophosphate levels it is predominantly nuclear, but it shifts to the cytosol under Pi-replete conditions (O’Neill et al, 1996; Chabert et al, 2023; Desfougères et al, 2016a; Auesukaree et al, 2004). Pho4-GFP localisation can hence serve as a readout for cytosolic Pi signalling. In wildtype cultures growing logarithmically on Pi-rich media, a small but significant fraction of the cells show nuclear accumulation of Pho4-GFP. Such activation of the PHO pathway in Pi-replete media can reflect a transient shortage of Pi during S-phase, when Pi utilization for biosynthesis may exceed the uptake capacity of the cell (Bru et al, 2016; Neef & Kladde, 2003; Pondugula et al, 2009). In addition, feedback signalling can maintain a fraction of these cells in a stable state of PHO pathway activation that had been triggered by a preceding period of transient Pi scarcity (Wykoff et al, 2007; Vardi et al, 2013, 2014).

We scored the fraction of cells showing this sign of Pi scarcity even in Pi-replete medium. This fraction was small (7%) for wildtype cells (Fig. 6). It increased fourfold upon hyperactivation of polyP synthesis, which we achieved by overexpression of the regulatory VTC subunit Vtc5 (Desfougères et al, 2016a). Ablation of polyP synthesis by deletion of the catalytic subunit Vtc4 had the opposite effect and reduced the frequency of cells with nuclear Pho4-GFP by half.

Impact of the vacuolar polyP/Pi cycle on cytosolic Pi signalling

A) Illustration of nucleo-cytoplasmic relocation of Pho4-GFP in response to Pi availability. Wildtype yeast cells were grown in SC medium under Pi replete conditions. During exponential phase (OD600nm=1), cells were transferred for 30 min to synthetic complete media with 200 µM phosphate (-Pi) or 7.5 mM Pi (+Pi) and imaged by fluorescence microscopy. B) The indicated yeast strains were logarithmically grown over night in SC medium with 7.5 mM Pi, harvested at OD600nm=1 and immediately imaged by fluorescence microscopy as in A. The graph shows the means and SEM of the percentage of cells showing Pho4 predominantly in the nucleus. n=3.

Deletion of the vacuolar polyphosphatases Ppn1 and Ppn2 prevents polyP turnover and leads to the accumulation of extremely long polyP chains (Gerasimaite & Mayer, 2017; Sethuraman et al, 2001). We may thus expect phosphate to remain fixed in the form of polyP instead of being made available for Pi reflux into the cytosol.

In line with this, the fraction of ppn11 ppn21 double mutants that showed nuclear Pho4-GFP was 3-fold higher than in wildtype. pho911 cells, in which we expect Pi export from the vacuoles to be impaired, showed two times higher frequency of nuclear Pho4-GFP than wildtype. pho911 vtc41 double mutants showed an even lower frequency of nuclear Pho4-GFP than wildtype, suggesting that the state of Pi starvation that pho911 favours is dependent on vacuolar polyP accumulation. Cells expressing pho91K237A as the sole source of Pho91 also showed a 50% lower frequency of nuclear Pho4-GFP. The pho91K237A allele generates an amino acid substitution in the InsPP-binding patch of the Pho91 SPX domain. It mimics the InsPP-free state (Wild et al, 2016) and hence low-Pi conditions (Chabert et al, 2023). Collectively, our results are consistent with the notion that loss of InsPP binding activates Pho91 to export Pi from vacuoles to the cytosol, enhancing repression of the PHO pathway.

Since InsPP levels in yeast decline in response to cellular phosphate availability (Lonetti et al, 2011; Chabert et al, 2023), we tested the impact of Pho91, Ppn1 and Ppn2 on the cellular levels of these metabolites using CE-MS, capillary electrophoresis coupled to mass spectrometry (CE-MS) (Qiu et al, 2020, 2023, 2021). The cells were harvested from the same Pi-replete growth conditions as for the microscopic assays above (Fig. 7).

Impact of Pho91 and vacuolar polyphosphatases on InsPP levels.

pho91τι and ppn1τιppn2τι cells, as well as their isogenic wildtypes, were logarithmically grown in Pi-replete SC medium as in Fig. 6. At OD600nm=1 the cells were extracted with perchloric acid as previously described (Wilson et al, 2015) and analyzed for the indicated InsPs through capillary electrophoresis coupled to mass spectrometry (CE-MS) as described (Qiu et al, 2023).

The ppn11 ppn21 double mutants showed a decrease by around one third in the inositol pyrophosphates 1,5-InsP8, 5-InsP7 and 1-InsP7, whereas InsP6 remained at a similar level as in wildtype. This is consistent with the 3-fold increase of cells with nuclear Pho4-GFP that these cells showed (Fig. 6). pho911 cells did not show significant changes for all four metabolites although they showed a partial shift of Pho4-GFP into the nucleus (Fig. 6). We attribute this discrepancy to the different nature of the assays. The microscopic assay for Pho4-GFP localization can pick up effects in a fraction of the cells because it offers single cell resolution. Inositol pyrophosphate analysis is an ensemble assay, in which changes in a smaller fraction of the population become diluted through the major pool that does not show the effect. For this reason, a change in inositol pyrophosphates affecting only 10% of the pho911 cells that show nuclear Pho4-GFP remains undetectable. The microscopic assay suggests that cytosolic Pi scarcity in ppn11 ppn21 affects more cells, perhaps because it is more profound, and it hence becomes detectable even in a whole-population-analysis.

Discussion

Our observations can be integrated with existing data on the properties of VTC (Wild et al, 2016; Hothorn et al, 2009; Müller et al, 2002; Gerasimaite et al, 2014, 2017; Liu et al, 2023; Guan et al, 2023; Pipercevic et al, 2023) and Pho91 (Hürlimann et al, 2007; Wang et al, 2015; Potapenko et al, 2018) to generate a working model explaining how an acidocalcisome-like organelle such as the yeast vacuole is set up to function as a Pi buffer for the cytosol. Under Pi-replete conditions, high InsPP levels activate VTC to polymerize Pi into polyP and translocate it into the vacuolar lumen. Here, the vacuolar polyphosphatases degrade polyP into Pi, filling the lumen with Pi (Sethuraman et al, 2001; Shi & Kornberg, 2005; Lichko et al, 2010; Gerasimaite & Mayer, 2017). Since the Pi exporter Pho91 is downregulated through InsPP binding to its SPX domain (Potapenko et al, 2018; Hürlimann et al, 2007; Wang et al, 2015), the Pi liberated through polyP hydrolysis accumulates in the vacuoles. Product inhibition of the polyphosphatases attenuates polyP hydrolysis once the vacuolar lumen has reached a Pi concentration above 30 mM. When the cells experience Pi scarcity, InsPP levels decline (Chabert et al, 2023). This activates Pho91 to release Pi from the vacuolar pool into the cytosol and stabilizes cytosolic Pi.

Working model of acidocalcisome-like vacuoles as Pi buffering systems.

ATP drives the conversion of Pi into polyP and its translocation into the organelle. Here, polyP is degraded by the vacuolar polyphosphatases Ppn1 and Ppn2 to establish a vacuolar pool of free Pi. Feedback inhibition of Pi gradually reduces polyP degradation, enabling the buildup of a vacuolar polyP stock. Cytosolic Pi scarcity decreases InsPP levels, which triggers two compensatory, SPX-controlled effects: The transfer of Pi from the cytosol into vacuoles through VTC ceases; and Pho91-dependent export of Pi from vacuoles is activated. Both measures synergize to stabilize cytosolic Pi.

Yeast cells do not only accumulate Pi as a rapidly accessible buffer for the cytosol. Under Pi-replete conditions they accumulate hundreds of millimolar of phosphate in the form of polyP (Urech et al, 1978). In contrast to the vacuolar reserve of Pi, which is presumed to be accessible immediately, mobilizing the polyP store takes minutes to hours (Nicolay et al, 1982; Pondugula et al, 2009; Bru et al, 2016). But the polyP store offers advantages in the form of high capacity - hundreds of millimolar of phosphate units can be stored in the form of polyP - and low osmotic activity of polyP (Dürr et al, 1979). Keeping such a large stock of a critical resource, which is often growth-limiting in nature, is relevant for the cells. In case of phosphate shortage, the vacuolar polyP store can be mobilized to enable the cells to complete the cell cycle and transition into G0 phase (Spain et al, 2015; Westenberg et al, 1989; Müller et al, 1992). This can consume substantial amounts of phosphate, because we can estimate that replicating the entire DNA (1.2*107 base pairs) immobilizes roughly 1 mM Pi in the cells; and cellular RNA is even 50 times more abundant than DNA (Warner, 1999), accounting for 50 mM phosphate. Phospholipids, which must also be synthesized to complete a cell cycle, fix phosphate in similar amounts (Lange & Heijnen, 2001). Thus, a large polyP store is necessary to guarantee that the cells can finish S-phase upon a shortage of phosphate sources. In accord with this notion, absence of the polyP store impairs cell cycle progression, nucleotide synthesis and induces genome instability (Bru et al, 2016, 2017). Also, a shift from non-fermentable carbon sources to fermentation of glucose leads to a strong requirement for Pi because the activation of glucose uptake and glycolysis depends on large amounts of phosphate-containing sugars and glycolytic intermediates (Nicolay et al, 1982, 1983; Gillies et al, 1981). Shortage of Pi restrains the abundance of these metabolites (Kim et al, 2023).

The properties of the regulatory circuit described above imply an inbuilt switch from vacuolar Pi accumulation to large-scale stocking of vacuolar polyP. Pi-replete conditions generate high cellular InsPP levels. These will not only reduce Pi efflux from the vacuoles through Pho91 and inactivate the vacuolar polyphosphatases, but at the same time stimulate continued polyP synthesis by VTC. Coincidence of these effects will favour storage and high accumulation of phosphate in the form of polyP. Conversely, depletion of the vacuolar Pi reservoir upon Pi scarcity in the medium will activate the vacuolar polyphosphatases. In combination with the downregulation of the polyP polymerase VTC through the decline of InsPPs this will mobilize the large vacuolar polyP reserve once the immediately available vacuolar Pi pool is gradually depleted.

The concentration of Pi inside vacuoles as a rapidly accessible Pi reserve, and the synthesis of a large polyP stock, come at an energetic cost because the transformation of Pi into polyP requires the formation of phosphoric anhydride bonds (Hothorn et al, 2009; Gerasimaite et al, 2014) and vacuolar Pi reaches 30 mM. This exceeds the cytosolic Pi concentration, which was measured through 31P-NMR in a variety of yeasts, yielding values of 5-17 mM (Nicolay et al, 1982, 1983). Cytosolic Pi can also be estimated based on data from several other studies (Auesukaree et al, 2004; Hürlimann et al, 2009; Theobald et al, 1996; Zhang et al, 2015; Pinson et al, 2004). Assuming that the cytosolic volume of a BY4741 yeast cell is 40 fL (Chabert et al, 2023), and 1 g of dry weight contains 40 * 109 yeast cells, these studies point to cytosolic values of 10-15 mM in Pi-replete media. Upon Pi starvation, this value rapidly drops up to fivefold, resulting in a strong Pi gradient across the vacuolar membrane (Okorokov et al, 1980; Shirahama et al, 1996). To replenish the cytosolic pool under Pi scarcity, Pho91 can exploit not only this Pi concentration gradient, but also the vacuolar electrochemical potential, which was shown to stimulate Pi export through the Pho91 homolog OsSPX-MFS3 from plant vacuoles (Wang et al, 2015).

Vacuolar Pi accumulation is driven indirectly through ATP in two ways. VTC uses ATP as a substrate and transfers the phosphoric anhydride bond of the ψ-phosphate onto a polyP chain (Hothorn et al, 2009). The growing polyP chain exits from the catalytic site directly towards the transmembrane part of VTC (Liu et al, 2023; Guan et al, 2023). This transmembrane part likely forms a controlled channel that can guide polyP through the membrane (Liu et al, 2023). Coupled synthesis and translocation require the V-ATPase (Gerasimaite et al, 2014), probably because polyP is highly negatively charged and therefore follows the electrochemical potential across the vacuolar membrane of 180 mV (inside positive) and 1.7 pH units (Kakinuma et al, 1981), which is generated through the proton pumping V-ATPase. Thus, the combination of the VTC complex and vacuolar polyphosphatases can be considered as a Pi pump that is driven by ATP through polyP synthesis and through the electrochemical potential for polyP translocation and Pi export.

It is likely that acidocalcisome-like organelles of other organisms act as buffers for cytosolic Pi similarly as described in our model for yeasts. This notion is supported by the conserved molecular setup of acidocalcisome-like organelles as well as by phenotypic similarities. The acidocalcisomes of Trypanosomes contain VTC, a Pho91 homolog and proton pumps in their membranes, and polyphosphatases in their lumen (Huang & Docampo, 2015; Billington et al, 2023; Fang et al, 2007; Lander et al, 2013; Ulrich et al, 2013; Scott et al, 1997). Also the acidocalcisome-like organelles of the alga Chlamydomonas contain such proteins and they accumulate polyP through VTC as a function of the availability of Pi, a proton gradient and metal ions (Zúñiga-Burgos et al, 2024; Blaby-Haas & Merchant, 2014; Long et al, 2023; Hong-Hermesdorf et al, 2014; Aksoy et al, 2014; Ruiz et al, 2001; Goodenough et al, 2019). Like in yeast, the polyP stores are mobilized upon Pi limitation (Sanz-Luque et al, 2020; Plouviez et al, 2021). We hence propose that acidocalcisome-like vacuoles may have a general role as feedback-controlled, rapidly accessible phosphate buffers for the cytosol, addressing a critical parameter for metabolism. However, given that acidocalcisome-like organelles accumulate not only phosphate but also multiple other metabolites and ions (Docampo, 2024), they are probably interlinked with cellular metabolism in multiple ways and might form an important hub for its homeostasis.

Materials and methods

Yeast strains and growth conditions

Saccharomyces cerevisiae cells were grown on yeast extract-peptone-dextrose (YPD: 1% yeast extract, 2% peptone and 2% dextrose) or in synthetic complete (SC) medium from Formedium, supplemented with sodium phosphate as needed). Yeast backgrounds used in this study were BY4741 and BY4742. Genetic manipulations of yeast were performed by homologous recombination according to published procedures and/or transformation with the indicated plasmids (Gietz & Schiestl, 2007; Güldener et al, 1996). Strains used are listed in Table 1.

PHO4 localization

Cells transformed with pRS415-PPHO4-PHO4-GFP, a plasmid expressing a PHO4-GFP fusion from the PHO4 promotor (Chabert et al, 2023), were grown exponentially overnight in SC medium without leucine (SC-Leu). Care was taken that the culture did not grow beyond a density of OD600=0.7. At this point, the cells were collected by brief centrifugation (15 sec, 3000 x g) in a tabletop centrifuge and resuspended in 1/10th to 1/20th of their own supernatant. Pho4-GFP localization was immediately checked by microscopy. An aliquot of the cells was washed twice with SC-Leu without Pi and then diluted in SC-Leu with 200 μM phosphate to OD600=0.7. After 30 minutes of incubation in this medium, Pho4-GFP localization was analysed by fluorescence microscopy on a LEICA DMI6000B inverted microscope equipped with a Hamamatsu ORCA-R2 (C10600-10B) camera, an XCite ® series 120Q UV lamp and a Leica 100x 1.4 NA lens.

Vacuole preparation

Vacuoles were purified from yeast cells essentially as described (D’Agostino & Mayer, 2019). Briefly, yeast cells were grown in 1L of YPD to an OD600 of 1.5. 330 ml of cells were collected by centrifugation and resuspended in 50 ml of 30 mM Tris-HCl pH 8.9, 10 mM DTT buffer. Suspensions were incubated in a 30° C water bath for 5 min and collected by centrifugation. The pellet was resuspended in 15 ml of spheroplasting solution (50 mM K-phosphate pH 7.5, 600 mM sorbitol in YPD with 0.2% D-glucose and 3600 U/ml lyticase) and incubated for 25 min at 30°C. Spheroplasts were collected by centrifugation (2500 x g, 3 min) and resuspended in 2.5 ml of 15% Ficoll 400 in PS buffer (10 mM PIPES-KOH pH 6.8, 200 mM sorbitol). 80 μg of DEAE-dextran were added under gentle mixing. After incubation on ice for 2 min and then at 30 °C for 80 sec, spheroplasts were transferred into Beckman SW41.1 tubes, overlaid with cushions of 8%, 4% and 0% Ficoll 400 in PS buffer, and centrifuged (150’000 x g, 90 min, 2 °C). Vacuoles were collected from the 0-4 % Ficoll interface. Their protein concentration was determined through Bradford assay using BSA as a standard.

Polyphosphatase activity of vacuolar lysates

Polyphosphatase activity was assayed as described previously (Gerasimaite & Mayer, 2017), with the following modifications. Isolated vacuoles were diluted to a final protein concentration of 0.002 mg/ml in 1 ml of reaction buffer (20 mM PIPES/KOH pH 6.8, 150 mM KCl, 1 mM ZnCl2 or MgCl2, 0.1% Triton X-100, 1xPIC, 1 mM PMSF and 300 µM polyP300) in the presence or absence of 30 mM of KH2PO4 and incubated at 27°C. At the indicated times, 80 µl aliquots were collected and the reaction was stopped by dilution with 160 µl of stop solution (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 12 mM EDTA pH 8.0, 0.1% Triton X-100, 15 µM DAPI). Remaining polyP was quantified by measuring polyP-DAPI fluorescence (λexc. 415 nm, λem. 550 nm) in a black 96 well plate in a Spectramax Gemini microplate fluorometer (Molecular Devices). A reaction containing boiled vacuoles was used as negative control.

Polyphosphate synthesis by isolated vacuoles

Polyphosphate synthesis was assayed as described (Gerasimaite et al, 2014). Isolated vacuoles were diluted to final protein concentration of 0.02 mg/ml on 1 ml of reaction buffer (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 0.5 mM MnCl2, 200 mM sorbitol) and the reaction was started by adding an ATP regenerating system (1 mM ATP-MgCl2, 40 mM creatine phosphate and 0.25 mg/ml creatine kinase). The mix was incubated at 27 °C. At different timepoints, 80 µl aliquots were mixed with 160 µl of stop solution (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 200 mM sorbitol, 12 mM EDTA, 0.15% Triton X-100 and 15 μM DAPI). PolyP synthesis was quantified through polyP-DAPI fluorescence (λexc. 415 nm, λem. 550 nm) in a black 96 well plate. A calibration curve was prepared using commercial polyP60 as a standard.

Phosphate quantification in isolated vacuoles

Isolated vacuoles were incubated as described for the polyP synthesis assay above. At different time points, 80 µl aliquots were centrifuged (3 min, 2000 x g, 2°C), the pellets were washed with 500 µl of washing solution (10 mM PIPES/KOH pH 6.8, 200 mM Sorbitol, 150 mM KCl) and centrifuged as before. The final pellet was resuspended with 100 µl of lysis buffer (10 mM PIPES/KOH pH 6.8, 200 mM sorbitol, 150 mM KCl, 12 mM EDTA, 0.1 % Triton). Free phosphate was quantified by adding 150 µl of molybdate-malachite green solution (1 mM malachite green, 10 mM ammonium molybdate,1M HCl) and reading the absorbance at 595 nm in a microplate photometer.

Inositol pyrophosphate synthesis, extraction and quantification

5-InsP7 was synthesized as described (Capolicchio et al, 2013; Wang et al, 2014). For quantification from cells, InsPPs extraction was performed as described previously (Kim et al, 2023), with the following modifications. Briefly, 3 ml of yeast cells at OD600nm=1 were collected using rapid vacuum-filtration on a polytetrafluoroethylene membrane filter (1.2 μm; Piper Filter GmbH, Germany). After snap freezing in liquid nitrogen, yeast cells on the membrane were resuspended in 400 μL of 1 M perchloric acid and lysed by bead beating (glass beads; 0.25-0.5 mm) for 10 min at 4 °C. After centrifugation at 13,000 rpm for 3 min at 4 °C, the supernatant was transferred into a new tube containing 3 mg of titanium dioxide (TiO2) beads (GL Sciences, Japan) which had been pre-washed twice with H2O and 1 M perchloric acid. The sample was gently rotated for 15 min at 4 °C. The TiO2 beads were collected by centrifugation at 13,000 rpm for 1 min at 4 °C and washed twice using 1 M perchloric acid. After the second washing step, the TiO2 beads were resuspended in 300 μL of 3 % (v/v) NH4OH and rotated gently at room temperature. After centrifugation at 13,000 rpm for 1 min, the eluants were transferred into a new tube and dried in SpeedVac (Labogene, Denmark) at 42 °C. InsPPs were measured through capillary electrophoresis coupled to mass spectrometry (CE-MS) as described (Qiu et al, 2021).

Quantification of the vacuolar proteome

Vacuoles were prepared as described above. During the purification procedure, a sample of the sedimented spheroplasts was withdrawn before DEAE dextran was added. These withdrawn spheroplasts constitute the “whole cell” extract. After the flotation step, the vacuoles withdrawn from the 4%-0% Ficoll interface were used as the vacuole fraction. For each fraction samples with 100 µg protein were precipitated by adding a final concentration 12.5% TCA for 10 min on ice. The proteins were sedimented (12’000 x g, 5 min, room temperature), the supernatant discarded, and the pellets were washed with ice-cold acetone and centrifuged as above twice. The final pellet was dried and dissolved in reducing SDS sample buffer (10% glycerol, 50 mM Tris-HCl pH 6.8, 2 mM DTT, 2% SDS, 0.002% bromophenol blue).

Samples dissolved in Laemmli buffer (50 mM Tris pH 6.8, 10 mM DTT, 2 % SDS, 0.1 % bromophenol blue, 10 % glycerol) were equalized in concentration by dilution in SP3 buffer (2% SDS, 10 mM DTT, 50 mM Tris, pH 7.5). 150 µg of protein per sample were heated at 95 °C for 5 min and cooled down. Reduced cysteine residues were alkylated by adding iodoacetamide (30 mM final) and incubating for 45 min at room temperature in the dark. Digestion was done by the SP3 method (Hughes et al, 2019) using magnetic Sera-Mag Speedbeads (Cytiva 45152105050250, 50 mg/ml). Beads were added at a ratio 10:1 (w:w) to samples, and proteins were precipitated on beads with ethanol (final concentration: 60 %). After 3 washes with 80% ethanol, beads were digested in 50 μl of 100 mM ammonium bicarbonate with 3.0 μg of trypsin (Promega #V5073). After 1 h of incubation at 37 °C, the same amount of trypsin was added to the samples for an additional 1 h of incubation. Supernatants were then recovered, transferred to new tubes, acidified with formic acid (0.5% final concentration), and dried by centrifugal evaporation. To remove traces of SDS, two sample volumes of isopropanol containing 1% TFA were added to the digests, and the samples were desalted on a strong cation exchange (SCX) plate (Oasis MCX; Waters Corp., Milford, MA) by centrifugation. After washing with isopropanol/1%TFA and 2% acetonitrile/0.1% FA, peptides were eluted in 200µl of 80% MeCN, 19% water, 1% (v/v) ammonia, and dried by centrifugal evaporation. Data-dependent LC-MS/MS analyses of samples were carried out on a Fusion Tribrid Orbitrap mass spectrometer (Thermo Fisher Scientific) interfaced through a nano-electrospray ion source to an Ultimate 3000 RSLCnano HPLC system (Dionex). Peptides were separated on a reversed-phase custom packed 45 cm C18 column (75 μm ID, 100Å, Reprosil Pur 1.9 μm particles, Dr. Maisch, Germany) with a 4-90% acetonitrile gradient in 0.1% formic acid at a flow rate of 250 nl/min (total time 140 min). Full MS survey scans were performed at 120’000 resolution. A data-dependent acquisition method controlled by Xcalibur software (Thermo Fisher Scientific) was used that optimized the number of precursors selected (“top speed”) of charge 2+ to 5+ while maintaining a fixed scan cycle of 0.6 s. Peptides were fragmented by higher energy collision dissociation (HCD) with a normalized energy of 32%. The precursor isolation window used was 1.6 Th, and the MS2 scans were done in the ion trap. The m/z of fragmented precursors was then dynamically excluded from selection during 60 s.

Data files were analysed with MaxQuant 1.6.14.0 (Cox & Mann, 2008) incorporating the Andromeda search engine (Cox et al, 2011). Cysteine carbamidomethylation was selected as fixed modification while methionine oxidation and protein N-terminal acetylation were specified as variable modifications. The sequence databases used for searching were the S. cerevisiae reference proteome based on the UniProt database (www.uniprot.org, version of June 6th, 2021, containing 6050 sequences), and a “contaminant” database containing the most usual environmental contaminants and enzymes used for digestion (keratins, trypsin, etc). Mass tolerance was 4.5 ppm on precursors (after recalibration) and 20 ppm on MS/MS fragments. Both peptide and protein identifications were filtered at 1% FDR relative to hits against a decoy database built by reversing protein sequences. The match between runs feature was enabled.

Data from the proteomic analysis have been deposited at the PRIDE database under the identifier PXD060102.

eLife Assessment

This study provides fundamental insights into eukaryotic phosphate homeostasis by demonstrating how yeast vacuoles dynamically regulate cytosolic phosphate levels. The conclusions are convincing, supported by an elegant combination of in vitro assays and in vivo measurements. This study will be of interest to cell biologists, particularly for those who are working in the field of phosphate metabolism.

Reviewer #1 (Public review):

The manuscript by Bru et al. focuses on the role of vacuoles as a phosphate buffering system for yeast cells. The authors describe here the crosstalk between the vacuole and the cytosol using a combination of in vitro analyses of vacuoles and in vivo assays. They show that the luminal polyphosphatases of the vacuole can hydrolyze polyphosphates to generate inorganic phosphate, yet they are inhibited by high concentrations. This balances the synthesis of polyphosphates against the inorganic phosphate pool. Their data further show that the Pho91 transporter provides a valve for the cytosol as it gets activated by a decline in inositol pyrophosphate levels. The authors thus demonstrate how the vacuole functions as a phosphate buffering system to maintain a constant cytosolic inorganic phosphate pool.

This is a very consistent and well-written manuscript with a number of convincing experiments, where the authors use isolated vacuoles and cellular read-out systems to demonstrate the interplay of polyphosphate synthesis, hydrolysis, and release. The beauty of this system the authors present is the clear correlation between product inhibition and the role of Pho91 as a valve to release Pi to the cytosol to replenish the cytosolic pool. I find the paper overall an excellent fit and only have a few issues, including :

(1) Figure 3: The authors use in their assays 1 mM ZnCl2 or 1mM MgCl2. Is this concentration in the range of the vacuolar luminal ion concentration? Did they also test the effect of Ca2+, as this ion is also highly concentrated in the lumen?

(2) Regarding the concentration of 30 mM K-PI, did the authors also use higher and lower concentrations? I agree that there is inhibition by 30 mM, but they cannot derive conclusions on the luminal concentration if they use just one in their assay. A titration is necessary here.

(3) What are the consequences on vacuole morphology if the cells lack Pho91?

(4) Discussion: The authors do not refer to the effect of calcium, even though I would expect that the levels of the counterion should affect the phosphate metabolism. I would appreciate it if they would extend their discussion accordingly.

(5) I would appreciate a brief discussion on how phosphate sensing and control are done in human cells. Do they use a similar lysosomal buffer system?

Reviewer #2 (Public review):

Summary:

This manuscript presents a well-conceived and concise study that significantly advances our understanding of polyphosphate (polyP) metabolism and its role in cytosolic phosphate (Pi) homeostasis in a model unicellular eukaryote. The authors provide evidence that yeast vacuoles function as dynamic regulatory buffers for Pi homeostasis, integrating polyP synthesis, storage, and hydrolysis in response to cellular metabolic demands. The work is methodologically sound and offers valuable insights into the conserved mechanisms of phosphate regulation across eukaryotes.

Strengths:

The results demonstrate that the vacuolar transporter chaperone (VTC) complex, in conjunction with luminal polyphosphatases (Ppn1/Ppn2) and the Pi exporter Pho91, establishes a finely tuned feedback system that balances cytosolic Pi levels. Under Pi-replete conditions, inositol pyrophosphates (InsPPs) promote polyP synthesis and storage while inhibiting polyP hydrolysis, leading to vacuolar Pi accumulation.

Conversely, Pi scarcity triggers InsPP depletion, activating Pho91-mediated Pi export and polyP mobilization to sustain cytosolic phosphate levels. This regulatory circuit ensures metabolic flexibility, particularly during critical processes such as glycolysis, nucleotide synthesis, and cell cycle progression, where phosphate demand fluctuates dramatically.

From my viewpoint, one of the most important findings is the demonstration that vacuoles act as a rapidly accessible Pi reservoir, capable of switching between storage (as polyP) and release (as free Pi) in response to metabolic cues. The energetic cost of polyP synthesis-driven by ATP and the vacuolar proton gradient-highlights the evolutionary importance of this buffering system. The study also draws parallels between yeast vacuoles and acidocalcisomes in other eukaryotes, such as Trypanosoma and Chlamydomonas, suggesting a conserved role for these organelles in phosphate homeostasis.

Weaknesses:

While the manuscript is highly insightful, referring to yeast vacuoles as "acidocalcisome-like" may warrant further discussion. Canonical acidocalcisomes are structurally and chemically distinct (e.g., electron-dense, in most cases spherical, and not routinely subjected to morphological changes, and enriched with specific ions), whereas yeast vacuoles have well-established roles beyond phosphate storage. A comment on this terminology could strengthen the comparative analysis and avoid potential confusion in the field.

Reviewer #3 (Public review):

Bru et al. investigated how inorganic phosphate (Pi) is buffered in cells using S. cerevisiae as a model. Pi is stored in cells in the form of polyphosphates in acidocalcisomes. In S. cerevisiae, the vacuole, which is the yeast lysosome, also fulfills the function of Pi storage organelle. Therefore, yeast is an ideal system to study Pi storage and mobilization.

They can recapitulate in their previously established system, using isolated yeast vacuoles, findings from their own and other groups. They integrate the available data and propose a working model of feedback loops to control the level of Pi on the cellular level.

This is a solid study, in which the biological significance of their findings is not entirely clear. The data analysis and statistical significance need to be improved and included, respectively. The manuscript would have benefited from rigorously testing the model, which would also have increased the impact of the study.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

The manuscript by Bru et al. focuses on the role of vacuoles as a phosphate buffering system for yeast cells. The authors describe here the crosstalk between the vacuole and the cytosol using a combination of in vitro analyses of vacuoles and in vivo assays. They show that the luminal polyphosphatases of the vacuole can hydrolyse polyphosphates to generate inorganic phosphate, yet they are inhibited by high

concentrations. This balances the synthesis of polyphosphates against the inorganic phosphate pool. Their data further show that the Pho91 transporter provides a valve for the cytosol as it gets activated by a decline in inositol pyrophosphate levels. The authors thus demonstrate how the vacuole functions as a phosphate buffering system to maintain a constant cytosolic inorganic phosphate pool.

This is a very consistent and well-written manuscript with a number of convincing experiments, where the authors use isolated vacuoles and cellular read-out systems to demonstrate the interplay of polyphosphate synthesis, hydrolysis, and release. The beauty of this system the authors present is the clear correlation between product inhibition and the role of Pho91 as a valve to release Pi to the cytosol to replenish the cytosolic pool. I find the paper overall an excellent fit and only have a few issues, including:

(1) Figure 3: The authors use in their assays 1 mM ZnCl2 or 1mM MgCl2. Is this concentration in the range of the vacuolar luminal ion concentration? Did they also test the effect of Ca2+, as this ion is also highly concentrated in the lumen?

The concentrations inside vacuoles reach those values. However, given that polyP can chelate divalent metal ions, what would matter are the concentrations of free Zn2+ or Mg2+ inside the organelle. These are not known. This is not critical since we use those two conditions only as a convenient tool to differentiate Ppn1 and Ppn2 activity in vitro. In our initial characterisation of Ppn2 (10.1242/jcs.201061), we had also tested Mn, Co, Ca, Ni, Cu. Only Zn and Co supported activity. Ca did not. Andreeva et al. (10.1016/j.biochi.2019.06.001) reached similar conclusions and extended our results.

(2) Regarding the concentration of 30 mM K-PI, did the authors also use higher and lower concentrations? I agree that there is inhibition by 30 mM, but they cannot derive conclusions on the luminal concentration if they use just one in their assay. A titration is necessary here.

The concentration of 30 mM was not chosen arbitrarily. It is the luminal Pi concentration that the vacuoles reached through polyP synthesis and hydrolysis when they entered a plateau of luminal Pi. We consider this as an upper limit because polyP kept increasing which luminal Pi did not. Thus, there is no physiological motivation for trying higher values. We have nevertheless added a titration to the revised version (new Fig. 3A).

(3) What are the consequences on vacuole morphology if the cells lack Pho91?

We had not observed significant abnormalities during a screen of the genome-wide deletion collection of yeast (10.1371/journal.pone.0054160), nor in other experiments with pho91 mutants, which we have not included in this manuscript due to a lack of effect.

(4) Discussion: The authors do not refer to the effect of calcium, even though I would expect that the levels of the counterion should affect the phosphate metabolism. I would appreciate it if they would extend their discussion accordingly.

The situation is much more complex because Ca2+ is not the only counterion. Major pools of counterions (up to hundreds of mM) are constituted by vacuolar lysine, arginine, polyamines, Mg, Zn etc. Their interplay with polyP is probably complex and worth to be treated in a dedicated project. If we wanted to limit the discussion of this complexity not to the simple statement that it is not understood, which is not very useful, we would have to engage in a lot of speculation. We feel that this would make the discussion lose focus and not contribute concrete insights.

(5) I would appreciate a brief discussion on how phosphate sensing and control are done in human cells. Do they use a similar lysosomal buffer system?

Mammalian cells have their Pi exporter XPR1 mainly on a lysosome-like compartment (10.1016/j.celrep.2024.114316). Whether and how it functions there for Pi export from the cytosol is not entirely clear. We have addressed this situation in the revised discussion section.

Reviewer #2 (Public review):

Summary:

This manuscript presents a well-conceived and concise study that significantly advances our understanding of polyphosphate (polyP) metabolism and its role in cytosolic phosphate (Pi) homeostasis in a model unicellular eukaryote. The authors provide evidence that yeast vacuoles function as dynamic regulatory buffers for Pi homeostasis, integrating polyP synthesis, storage, and hydrolysis in response to cellular metabolic demands. The work is methodologically sound and offers valuable insights into the conserved mechanisms of phosphate regulation across eukaryotes.

Strengths:

The results demonstrate that the vacuolar transporter chaperone (VTC) complex, in conjunction with luminal polyphosphatases (Ppn1/Ppn2) and the Pi exporter Pho91, establishes a finely tuned feedback system that balances cytosolic Pi levels. Under Pi-replete conditions, inositol pyrophosphates (InsPPs) promote polyP synthesis and storage while inhibiting polyP hydrolysis, leading to vacuolar Pi accumulation.

Conversely, Pi scarcity triggers InsPP depletion, activating Pho91-mediated Pi export and polyP mobilization to sustain cytosolic phosphate levels. This regulatory circuit ensures metabolic flexibility, particularly during critical processes such as glycolysis, nucleotide synthesis, and cell cycle progression, where phosphate demand fluctuates dramatically.

From my viewpoint, one of the most important findings is the demonstration that vacuoles act as a rapidly accessible Pi reservoir, capable of switching between storage (as polyP) and release (as free Pi) in response to metabolic cues. The energetic cost of polyP synthesis-driven by ATP and the vacuolar proton gradient-highlights the evolutionary importance of this buffering system. The study also draws parallels between yeast vacuoles and acidocalcisomes in other eukaryotes, such as Trypanosoma and Chlamydomonas, suggesting a conserved role for these organelles in phosphate homeostasis.

Weaknesses:

While the manuscript is highly insightful, referring to yeast vacuoles as "acidocalcisome-like" may warrant further discussion. Canonical acidocalcisomes are structurally and chemically distinct (e.g., electrondense, in most cases spherical, and not routinely subjected to morphological changes, and enriched with specific ions), whereas yeast vacuoles have well-established roles beyond phosphate storage. A comment on this terminology could strengthen the comparative analysis and avoid potential confusion in the field.

Yeast vacuoles show all major chemical features of acidocalcisomes. They are acidified, contain high concentrations of Ca, polyP (which make them electron-dense, too), other divalent ions, such as Mg, Zn, Mn etc, and high concentrations of basic amino acids. Thus, they clearly have an acidocalcisome-like character. In addition, they have hydrolytic, lysosomelike functions and, depending on the strain background, they can be larger than acidocalcisomes described e.g. in protists. We have elaborated on this point in the introduction of the revised version.

Reviewer #3 (Public review):

Bru et al. investigated how inorganic phosphate (Pi) is buffered in cells using S. cerevisiae as a model. Pi is stored in cells in the form of polyphosphates in acidocalcisomes. In S. cerevisiae, the vacuole, which is the yeast lysosome, also fulfills the function of Pi storage organelle. Therefore, yeast is an ideal system to study Pi storage and mobilization.

They can recapitulate in their previously established system, using isolated yeast vacuoles, findings from their own and other groups. They integrate the available data and propose a working model of feedback loops to control the level of Pi on the cellular level.

This is a solid study, in which the biological significance of their findings is not entirely clear. The data analysis and statistical significance need to be improved and included, respectively. The manuscript would have benefited from rigorously testing the model, which would also have increased the impact of the study.

It is not clear to us what the reviewer would see as a more rigorous test of the model.

Reviewer #1 (Recommendations for the authors):

(1) Figure 2: Why do the authors label the blue curve in A and B as BY and in C and D as WT? Is this a different genetic background they used here? This should be specified in the legend.

No, it is the same background. The figures had been reshuffled before submission and we overlooked to replace "BY" by "WT". This has been corrected. Now we consistently use WT in all figures

(2) Figure 4 has different scaling for the two panels, which should be labeled as A and B. I am aware that the authors do this for comparison, but it is rather confusing at first glance. I recommend having them at the same scale.

We chose this representation on two separate scales because this figure shall primarily illustrate that the shift between pho91 and WT curves vanishes in the presence of IP7. We now highlight in the figure legend that the scales are different to avoid confusion.

(3) Figure 8: I would appreciate a model with normal and low Pi concentrations in comparison, as this is what the authors worked out.

We have modified the figure. It now compares Pi-rich and Pi-limited scenarios.

(4) Minor issue: Wouldn't it make more sense to show the molar concentration in the Figures rather than the nmol of Pi/ug of protein? I am aware that this would require information on the vacuole volume rather than the reaction volume, and the authors do this calculation later on.

It depends. We often chose this representation because it illustrates the price to pay (metabolic input in terms of protein that must be dedicated to this task) to sequester a certain quantity of Pi. But, as we provide the corresponding Pi concentration in the text, this information is accessible to the reader, too.

Reviewer #2 (Recommendations for the authors):

As stated above in the weaknesses section, while functional parallels exist, canonical acidocalcisomes are structurally and chemically distinct, typically smaller, electron-dense, and enriched with cations. Whereas yeast vacuoles are larger, multifunctional organelles with well-established roles beyond phosphate storage. Explicitly addressing these differences would strengthen the comparative framework and prevent potential confusion in interpreting the evolutionary relationships between these organelles.

We agree to some degree, which is the reason why we refer to vacuoles as acidocalcisome-like organelles. In fact, vacuoles share virtually all defining chemical traits of acidocalcisomes. They just have a second functional domain as hydrolytic, lysosome-like organelles. Given the plasticity of endo-lysosomal compartments, and acidocalcisomes belong to this group because of their biogenesis through the AP3 pathway, this is not shocking to us. But the reviewer's comment made us realize that it is better to explicitly address this point. We have added a section to the introduction to do this.

Reviewer #3 (Recommendations for the authors):

(1) Page 8: It is unclear why the authors only estimated the Pi concentration in wild-type vacuoles. This should also be done for vacuoles from other strains.

This information is inherent in Figure 2. PolyP hyperaccumulating strains show the same plateau as the wildtype, meaning that they also reach around 30 mM luminal Pi concentration, whereas vtc4 vacuoles reach only around 1/10th of that increase, indicating that they remain at 3 mM. We mention this now in the text.

(2) The attempts of the localization of Pho91 through tagging are not satisfactory. The author described different localizations for Pho91 depending on whether it was tagged on the N- or C-terminus or when Nterminally tagged and overexpressed using two strong promoters. While it is not uncommon that proteins show different localization patterns, depending on where the tag is inserted, it is possible that one of the tags would reflect the localization of the endogenous protein. There is an easy way to test this, in particular when Pho91 is endogenously tagged. pho91∆ has reported phenotypes such as abnormal vacuolar morphology or increased autophagy. They could also measure PI content in vacuoles. The authors could compare the phenotypes of the endogenously tagged strains with WT and a pho91∆ strain.

Indeed, the attempts to localise the protein through fluorescent tags are unsatisfactory, in our hands as in the hands of others. We would not have created a series of many different tagged versions (we present only a selection of these in the manuscript) if the creation of a faithful reporter for Pho91 localisation were so straightforward. Expression from the endogenous promoter yields quite low signals (which is why others have overexpressed their GFP fusion from strong promotors). But overexpression brings at least a significant part of the protein to the cell surface, where it can then function as Pi importer and suffice to restore much of the maximal Pi uptake capacity that genuine plasma membrane transporters provide and support normal growth of the cells (Wykoff & O’Shea, 2001). But the localisation pattern of Pho91-GFP, likewise overexpressed from a strong promotor, does not reflect this plasma membrane localisation (see the references that the reviewer mentioned under (3)). The published overexpressed GFP-fusions localise only to the vacuole, suggesting that even in this case the GFP tag may create an artefact. Therefore, we went through a large variety of Pho91 gene fusions, which led us to the conclusion that the protein is very sensitive to tags at both ends and that fusion proteins hence are unlikely to reliably report the correct location of the protein. Given this, we resorted to quantitative proteomics to clarify the issue. This quantitative experiment goes beyond previously published proteomics analyses that the reviewer mentions under (3), which found the protein in the vacuolar fraction but did not calculate the enrichment factors, which is crucial.

A strong phenotype of abnormal vacuolar morphology is not apparent in our cultures.

(3) Moreover, Pho91 has been identified as a component enriched in vacuolar-mitochondria contact sites (vCLAMP), and this localization was confirmed with GFP-Pho91 (PMID: 25026036). Likewise, PMID: 35175277 also detected Pho91 by mass spectrometry as a vacuolar protein and showed endogenously tagged GFP-Pho91 on the vacuole (co-staining with Vph1). The authors may request the strains from the authors of these papers and use them for their experiments. PMID: 17804816, the oldest of the three reports (from 2007) reports a GFP-Pho91 under either TEF or ADH promoter that localizes to the vacuole. They also showed that the fusion protein is functional. These and other experiments led them to conclude that Pho91 exports phosphate from the vacuolar lumen to the cytoplasma.

We have now included these references. As argued above, we have analysed also the strains from PMID17804816. The observed clear localisation of the fusion protein to vacuoles is only visible upon overexpression, not upon expression from the endogenous locus. Apparently also this construct is unlikely to report Pho91 localisation reliably (though, by chance, overexpression leads it to the correct location). Thus, we maintain our conclusion that C- or N-terminally GFP-tagged versions of Pho91 are unreliable tools for localising the protein.

(4) The impact of pho91∆ on Pho4-GFP nuclear localization is modest at best (increase from 5% of cells showing Pho4-GFP in the nucleus in WT vs 10% in pho91∆), and only somewhat stronger in ppn1∆/ppn2∆. This means 90% of pho91∆ cells do not respond, and Pho4-GFP stays cytoplasmic. It is unclear how the author can derive a meaningful conclusion from these data. Moreover, are these data really supporting the model, or do these data rather indicate that there are additional factors/pathways needed? What is the biological significance of the marginal increase from 5% to 10% of cells that would respond? What happens to the cells that cannot respond? Will they die or at least have a growth disadvantage? It would be useful to provide some functional studies.

We should have explained the nature of the assay better. The experiment exploits the fact that dividing yeast cells transiently fall into a state of Pi scarcity during S-phase. Since S-phase is less than a quarter of the cell cycle, only a small fraction of the cells transiently activates the PHO pathway. These cannot be well characterised by ensemble assays, but microscopy circumvents this background of the whole population and picks them up very clearly, allowing to quantify them. We have adapted the respective chapter in the results section to improve the description of this experiment.

(5) The quantification of the data is suboptimal, as in most assays the mean and standard error of the mean (SEM) are given. SEM is not really appropriate in these cases because it gives only the error of the mean and not of the entire data. Therefore, the standard deviation (SD) is needed, which reports on the variability of the data, and which is usually much larger than the SEM. Using the SD, would also allow the authors to do proper statistical analysis, which is missing entirely in this manuscript.

SEM also comprises the variability of the data. It is linked with the SD (SEM=SD/SQRT(n)), but SEM also considers the number of the experiments n. The main goal is to compare the means, and SEM is an appropriate and frequently used tool for this because it illustrates how well the arithmetic mean may estimate the true mean of the population. Therefore, we kept the SEM but have added tests of significance for the differences shown.

(6) Statistical testing in Figure 7 is essential as the effects are very small. Again, are these changes big enough for a biologically meaningful response? The authors should at least discuss this.

Our previous time course analyses of InsPP dynamics, performed under comparable conditions as in this study, showed that InsP8 decreases by around 50% in the first 30 min after transfer to Pi starvation (DOI: https://doi.org/10.7554/eLife.87956) and that this decline is already sufficient to trigger the PHO starvation program, as assessed by Pho4-GFP translocation into the nucleus. Thus, a 50% decrease, which is observed in ppn1 ppn2 mutants, is functionally significant. We have now also evaluated statistical significance in Fig. 7, which is given for the 50% reduction of InsP8 and 1-InsP7 in ppn1 ppn2.

Minor points:

(1) There are a number of smaller edits (use of italic or better the absence thereof, lacking information in the reference list, and some typos).

Thank you. We have corrected those.

(2) The exact n should be given in the Figure legend.

Corrected.

(3) Page 8, line 8: it would be nice to have a picture of the wild-type vacuoles and what you measured.

We now present a sample image in the new Suppl. Fig. 1.

(4) PMID: 11779791 showed already that Pho91 cannot rescue the absence of the plasma membrane Pi transporters. This study should be at least cited.

This is not quite correct. The study that the reviewer mentions showed that Pho91 supports slower growth and the authors concluded that "A synthetic lethal phenotype was observed when (all) five phosphate transporters were inactivated...". We had cited the same group and the same first author, just using their later study (Wykoff et al., 2007) that had recapitulated the results from PMID11779791 and showed in addition quite good growth of the PHO91 expressing strain on YPD (Suppl. Fig. 2). We had obtained the strains from this group. In reproducing their experiments, we noticed that the growth of Pho91 that these authors had observed is due to incomplete repression of Pho84. They had overexpressed Pho84 from a galactose inducible promotor to generate a background with a regulatable Pi transporter. This trick allowed them to conveniently manipulate the strain and reduce (but not abolish) Pho84 expression by transferring the cells from galactose to glucose for their experiments. Therefore, we chose a more rigorous plasmid shuffling strategy to test the individual Pi transporter, which allows an assessment without the leaky background expression of Pho84 on glucose. In contrast to O'Shea and colleagues, we observed zero growth of a strain expressing only PHO91. We have revised the results section to make this discrepancy more evident and provide a better motivation for our experiment.

(5) It would be nice to see the actual data in Figure 6; not only a quantification.

We illustrate the phenotype of nuclear Pho4-GFP in panel A. Showing all the images necessary to appreciate the differences between the strains would require including many dozens of images into the figure, which would not be useful.

Version 2

Abstract

Cells experience strong variations in the consumption and availability of inorganic phosphate (Pi). Since Pi is an essential macronutrient but excess Pi has negative impacts on nucleotide hydrolysis and metabolism, its concentration must be maintained in a suitable range. Conserved storage organelles, acidocalcisomes, provide this buffering function. We used acidocalcisome-like yeast vacuoles to study how such organelles are set up to for this task. Our combined in vitro and in vivo analyses revealed that their ATP-driven polyphosphate polymerase VTC converts cytosolic Pi into inorganic polyphosphates (polyP), which it transfers into the vacuole lumen. Luminal polyphosphatases immediately hydrolyse this polyP to establish a growing reservoir of vacuolar Pi. Product inhibition by this Pi pool silences the polyphosphatases, caps Pi accumulation, and favours vacuolar polyP storage. Upon cytosolic Pi scarcity, the declining inositol pyrophosphate levels activate the vacuolar Pi exporter Pho91 to replenish cytosolic Pi. In this way, acidocalcisome-like vacuoles constitute a feedback-regulated buffering system for cytosolic Pi, which the cells can switch between Pi accumulation, Pi release, and high-capacity phosphate storage through polyP.

Introduction

Cells actively manage the concentration of Pi in their cytosol because they must strike a balance between conflicting goals (Austin & Mayer, 2020). On the one hand, Pi is a product of nucleotide-hydrolysing reactions. Its concentration has a significant impact on the free energy that these reactions can provide to drive metabolism. Since, for his reason, excessive Pi concentrations might stall metabolism, we can expect that cytosolic Pi must remain limited. But Pi is also an essential macronutrient. As a major constituent of nucleic acids and phospholipids, and as an important modifier of proteins, carbohydrates and many metabolites, it is consumed in large quantities for anabolic reactions. This can lead to situations where progression of the S-phase of the cell cycle is limited by the Pi uptake capacity of the cells (Bru et al, 2016, 2017; Gillies et al, 1981; Neef & Kladde, 2003). Rapidly dividing cells, such as yeast cells, remedy this problem by maintaining phosphate stores in the form of inorganic polyP. PolyPs are chains of Pi linked through phosphoric anhydride bonds, which can be stored in acidocalcisome-like organelles (Urech et al, 1978; Okorokov et al, 1980). Acidocalcisomes are lysosome-related organelles that have been characterised by pioneering work mainly in Trypanosomes, but they have been isolated from a multitude of other organisms (Docampo et al, 2005). Their defining chemical features are their acidity, high concentrations of polyP, Ca, Mg, Zn, Fe, K, Na, and basic amino acids and they are implicated in osmoregulation and the homeostasis of phosphate and metals (Docampo et al, 2005; Girard-Dias et al, 2023; Docampo, 2024). Yeast vacuoles share all these properties and are hence good and easily accessible models for studying aspects of acidocalcisomal mechanisms, such as the synthesis and turnover of polyP (Dürr et al, 1979; Huber-Wälchli & Wiemken, 1979; Urech et al, 1978; Austin & Mayer, 2020). But they have an additional, lysosome-like face, enabling them to assume also hydrolytic degradative functions for the cells (Li & Kane, 2009). Therefore, we refer to them as acidocalcisome-like organelles.

The polyP stores of vacuoles are accessed by cells when Pi becomes limiting in the environment, or when they face a sudden enhanced Pi consumption, e.g. during metabolic transitions from respiratory growth to fermentation, which requires much more phosphorylated metabolites (Shirahama et al, 1996; Thomas & O’Shea, 2005; Gillies et al, 1981; Nicolay et al, 1982, 1983). In such cases, cellular polyphosphate content decreases. It has therefore been proposed that polyPs are re-converted into Pi to buffer shortages in cytosolic Pi (Nicolay et al, 1983, 1982). While this hypothesis is straightforward, we are lacking a coherent concept of how a Pi buffer based on an acidocalcisome-like organelle might work. Pioneering work on these organelles has revealed several characteristic and conserved features, such as their acidity, and their high content of basic amino acids and divalent cations (Docampo & Huang, 2016; Docampo, 2024). Of direct relevance to Pi homeostasis is the high capacity of acidocalcisome-like organelles for storing polyP. PolyP can greatly vary in length, from two to hundreds of phosphate units. The membrane of acidocalcisome-like organelles can carry polyP polymerases, such as VTC, and Pi transporters, such as Pho91 (Jimenez & Docampo, 2015; Hürlimann et al, 2007; Wang et al, 2015; Huang & Docampo, 2015; Gerasimaite & Mayer, 2016; Müller et al, 2003, 2002). VTC is a coupled polyP polymerase and translocase (Gerasimaite et al, 2014), which synthesizes polyP by transferring the ψ-phosphate of cytosolic ATP onto an elongating polyP chain (Hothorn et al, 2009). PolyP is generated by a catalytic domain in the centre of the Vtc4 subunit of the complex (Hothorn et al, 2009). The activity of this subunit is controlled through the SPX domains of VTC (Wild et al, 2016), which may associate into a dimerized inactive state (Pipercevic et al, 2023). They can be released from this state through the action of inositol pyrophosphates, signalling molecules that accumulate when Pi is abundant in the cytosol and activate Pi storage in the form of polyP. The Pi state of the cytosol is communicated to VTC through a specific inositol pyrophosphate, 1,5-IP8 (Gerasimaite et al, 2017; Chabert et al, 2023).

The postulated direct channelling of polyP from the site of its synthesis through the vacuolar membrane (Gerasimaite et al, 2014) is facilitated by the structure of the VTC complex. The catalytic domain of VTC has its exit for polyP right at the entry of a polyP-conducting channel, which is formed through the transmembrane domains of the VTC complex itself (Liu et al, 2023; Guan et al, 2023; Müller et al, 2002, 2003; Gerasimaite et al, 2014). This channel was proposed to exist in an open and closed conformation and to be gated through polyP (Liu et al, 2023). PolyP translocation may be driven by the electrochemical potential across the membrane, which could move the highly negatively charged polyP chain through electrophoresis. This can explain why efficient polyP synthesis depends on proton pumps such as the V-ATPase or H+-pumping pyrophosphatases (Gerasimaite & Mayer, 2016; Gerasimaite et al, 2014; Freimoser et al, 2006; Lemercier et al, 2004).

The transporter Pho91 was proposed to export vacuolar Pi (Hürlimann et al, 2007). Several properties of yeast Pho91 and its homologs from other organisms are consistent with this view. Its ablation increases vacuolar Pi and polyP content, in yeast as well as in trypanosomes (Hürlimann et al, 2007; Jimenez & Docampo, 2015; Farofonova et al, 2023), and it weakly induces the phosphate starvation program in yeast (Pinson et al, 2004), suggesting that it may induce cytosolic Pi scarcity. Furthermore, patch-clamp analyses of Pho91-like channels from yeast, trypanosomes and plants have shown their Pi permeability and the dependence of their directionality on a pH gradient across the membrane (Potapenko et al, 2018, 2019; Wang et al, 2015). In the presence of a pH gradient, the Pho91 homolog from rice, OsSpx-MFS3, mediates Pi flux along the proton gradient, consistent with a function in exporting Pi from the acidic lumen of vacuoles (Wang et al, 2015) towards the cytosol.

Acidocalcisome-like organelles also contain polyphosphatases in their lumen (Lander et al, 2016; Gerasimaite & Mayer, 2016; Kulakovskaya et al, 2021; McCarthy & Downey, 2023), which can convert polyP back into Pi. In baker’s yeast, two vacuolar polyphosphatases are known, Ppn1 and Ppn2 (Gerasimaite & Mayer, 2017; Sethuraman et al, 2001; Andreeva et al, 2019). This means that a chain of polyP, when being synthesized by VTC and arriving in the vacuolar lumen, is immediately exposed to hydrolytic enzymes that will degrade it. While this seems at first sight paradoxical, we explored the hypothesis that the co-existence of polyP-synthesizing and polyP-hydrolysing activities might be a key feature conveying to acidocalcisome-like organelles the capacity to buffer cytosolic Pi. That these organelles have a critical role to play in this process is illustrated by observations in yeast, where artificial up- or down-regulation of vacuolar polyP synthesis suffices to drive the cytosol into a state of Pi starvation or Pi excess, respectively (Desfougères et al, 2016a). Furthermore, the presence of polyP reserves delays the activation of the transcriptional phosphate starvation response, the PHO pathway (Thomas & O’Shea, 2005). We hence explored the capacity of isolated yeast vacuoles to interconvert polyP and Pi, and we characterized the roles played by the vacuolar polyphosphatases Ppn1 and Ppn2 and the vacuolar Pi transporter Pho91. Our observations can be combined with previous findings to yield a coherent model of how an acidocalcisome-like organelle can operate as a Pi buffer for the cytosol.

Results

We explored the interplay of VTC, polyphosphatases and Pho91 in the accumulation of polyP and Pi inside vacuoles using an in vitro system with purified organelles. Vacuoles can be isolated in intact form when the cells are gently opened by enzymatic digestion of the cell wall and disruption of the cell membrane by low concentrations of DEAE-dextran (Dürr et al, 1975). Organelles isolated in this way can perform many of their normal cellular functions, such as membrane fusion, membrane fission and autophagy (Michaillat et al, 2012; Sattler & Mayer, 2000; D’Agostino & Mayer, 2019; Kunz et al, 2004). They also contain active polyP polymerase (VTC) and active polyphosphatases (Ppn1 and Ppn2) and they can synthesize and import polyP (Gerasimaite et al, 2014; Gerasimaite & Mayer, 2017; Gerasimaite et al, 2017).

Vacuolar Pi accumulation depends on polyP synthesis

Purified vacuoles were incubated with an ATP-regenerating system under conditions that allow these organelles to synthesize polyP in vitro (Gerasimaite et al, 2014). After different periods of incubation, the organelles were sedimented and solubilized in detergent. Vacuolar accumulation of polyP was assayed through DAPI fluorescence and Pi was measured through malachite green. VTC is stimulated by a variety of inositol pyrophosphates, most efficiently by 1,5-InsP8 (Gerasimaite et al, 2014, 2017). Since 1,5-InsP8 is not commercially available, we used saturating concentrations of 5-InsP7 for our experiments (Pavlovic et al, 2015). This compound stimulates VTC with a higher EC50 but to the same maximal activity as 1,5-InsP8, and it was more readily available to us for routine experiments. Under the conditions used here, the organelles rapidly and efficiently produced polyP (Fig. 1), in line with previous results (Gerasimaite et al, 2014, 2017). This production was stimulated by 5-InsP7, reaching 0.7 nmol of phosphate units per µg of vacuolar protein and min. This means that already within 10 min, wildtype vacuoles produced a mass of polyP equivalent to their total protein content, indicating how efficiently these organelles synthesize polyP. This signal was entirely dependent on VTC, because it was not observed in a mutant lacking the catalytic subunit of VTC (vtc4Δ).

VTC- and 5-IP7-dependent polyP synthesis by isolated vacuoles

Vacuoles were isolated from logarithmic cultures of BY4742 wildtype cells (WT) or from isogenic vtc4Δ cells strains. They were incubated in polyphosphate synthesis assays without (A) or (B) in presence of 50 µM 5-InsP7. At indicated times, aliquots were withdrawn, solubilized in Triton X-100 and polyphosphate was quantified through the polyP-dependent fluorescence of added DAPI. Means ± SD (standard deviation) of three independent experiments are shown. * p<0.05, ** p<0.005 for a paired t-test comparing WT with and without 5-InsP7 at each time point.

Isolated wildtype vacuoles also accumulated Pi with significant efficiency, at an initial rate of at least 0.3 nmol µg-1 min-1 (Fig. 2A). Vacuoles from vtc4Δ cells showed only 8% of this Pi signal. Most of it was not time-dependent and hence may represent a background signal from the organelle preparation. In sum, vacuolar Pi accumulation depended on polyP synthesis through VTC.

VTC-dependent accumulation of Pi in isolated vacuoles

Vacuoles were isolated from the indicated, logarithmically growing BY4742 (WT) cells. PolyP overproduction was achieved either through (A, B) overexpressing VTC5 from the strong GPD promotor, or (C,D) by expressing the hyperactivating vtc3k126A and vtc4K129A alleles form their native promotors as the sole source of these two proteins. The vacuoles were incubated with an ATP-regenerating system and 50 µM 5-InsP7 under conditions allowing polyP synthesis. At indicated times, the vacuoles were solubilized with Triton X-100 and Pi was assayed through malachite green. Graphs represent the mean ± SEM of three independent experiments for each strain. * p<0.05, ** p<0.005, *** p<0.001 from a paired t-test comparing each time WT vs VTC5 overexpression or vtc3k126A vtc4K129A.

The kinetics of polyP and Pi accumulation were remarkably different. Whereas polyP continued to accumulate throughout the entire incubation period, Pi accumulated rapidly in the initial phase but reached a plateau within 30 min (Fig. 2A, B). Thus, although polyP production was essential for Pi accumulation, it could not become the limiting factor for it in the second phase. This is further illustrated by Pi accumulation in strains showing enhanced polyP synthesis, such as overexpressors of VTC5 (Fig. 2A, B) or strains carrying hyperactivating substitutions in the SPX domains of VTC (vtc3K126A vtc4K129A; Fig. 2 C, D) (Desfougères et al, 2016b; Wild et al, 2016).

Vacuoles from these strains accumulated polyP at a two to seven times higher initial rate than the wildtype and to higher concentrations. Nevertheless, their accumulation of Pi arrested at the same level as that of wildtype vacuoles, at 5 nmol µg-1 vacuolar protein. To estimate the corresponding luminal concentration of Pi, we measured the diameters of 100 isolated wildtype vacuoles and counted the number of vacuoles per µg of vacuolar protein. With an average diameter of 0.8 µm (Suppl. Fig. 1) and 5*107 vacuoles/µg of protein we can estimate that Pi accumulation in vacuoles incubated with 5-InsP7 reached a plateau at a luminal concentration of around 30 mM. vtc4Δ vacuoles showed only a very small increase of less than 3 mM in luminal Pi content.

The 30 mM of Pi accumulating in wildtype vacuoles is a substantial concentration, which is in the range of the in vivo Pi concentration of 25 mM that was measured by 31P-NMR spectroscopy of yeast cells under conditions where vacuolar Pi dominates the signal (Okorokov et al, 1980). Since, upon Pi scarcity, cytosolic Pi drops to 1 mM (Okorokov et al, 1980), a substantial Pi concentration gradient across the vacuolar membrane could drive rapid replenishment of the cytosol from a readily accessible vacuolar pool of Pi.

Vacuolar Pi accumulation is limited by feedback inhibition of Ppn1 and Ppn2

We asked why vacuolar Pi accumulation quickly forms a plateau whereas vacuolar polyP continues to accumulate. To this end, we explored the hypothesis that vacuolar polyphosphatases become inhibited when their product, Pi, has accumulated. To test polyphosphatase activity we liberated the polyphosphatases from isolated vacuoles by detergent lysis, followed by incubation with synthetic polyP as a substrate and DAPI as an indicator for polyP. PolyP was efficiently degraded by this extract. The apparent polyPase activity was significantly reduced by 3 mM Pi and efficiently silenced by 30 mM Pi (Fig. 3A). To differentiate the contributions of Ppn1 and Ppn2, we also analysed the respective deletion mutants and performed the assay at higher polyP concentration in the presence Mg2+, which favours Ppn1 activity, or in the presence of Zn2+, which supports activity of Ppn1 and Ppn2 (Gerasimaite & Mayer, 2017). In the presence of Zn2+, the substrate was consumed in less than 3 min (Fig. 3). Degradation was delayed in vacuoles from ppn1Δ or ppn2Δ mutants, and it was suppressed in vacuoles from a ppn1Δ ppn2Δ mutant, in which both polyphosphatases were ablated. In incubations with only Mg2+ instead of Zn2+, which stimulates the activity of Ppn1 much more than that of Ppn2, polyP degradation was slower and genetic ablation of Ppn1 sufficed to stabilize polyP. The addition of 30 mM potassium phosphate, which is equivalent to the maximal Pi accumulation in vacuoles that we observed in Fig. 2, attenuated degradation of polyP 5-to 10-fold, both through Ppn1 (assayed in ppn2Δ and in samples without Zn2+) and through Ppn2 (assayed in ppn1Δ). That both polyphosphatases are inhibited by Pi at this concentration is consistent with the notion that product inhibition of Ppn1 and Ppn2 might limit the conversion of polyP into Pi in the vacuolar lumen and thus define the maximal concentration of the vacuolar Pi reservoir.

Effect of Pi on polyphosphatase activity.

A. Pi titration. Vacuoles were isolated from logarithmically growing wildtype cells (BY4742). The organelles were diluted in polyphosphatase assay reaction buffer, which contained 0.1% Triton X-100 and hence liberated the luminal polyphosphatases. This lysate was incubated with 30 μM polyP300 as a substrate and supplemented with 1 mM ZnCl2 and the indicated concentrations of K-Pi pH 6.8. After the indicated times of incubation, the remaining polyP was quantified through DAPI. The DAPI signal at the beginning of the incubation served as 100% reference. Graphs represent the means ± SEM of three independent experiments. B, C. Differentiation of Pi effects on Ppn1 and Ppn2. Vacuoles were isolated from the indicated, logarithmically growing strains, lysed and used in polyP degradation assays with 300 µM polyP300 as in (A). These assays were performed in the presence or absence of 30 mM K-Pi pH 6.8 and 1 mM MgCl2 (B) or, instead of this, with 1 mM ZnCl2 (C) as cation supporting catalytic activity.

Pho91 limits vacuolar Pi accumulation in an inositol pyrophosphate-dependent manner

To allow vacuoles to function as a Pi buffer for the cytosol, the Pi in the vacuolar lumen should become accessible in a regulated manner. The Pi transporter Pho91 is a prime candidate for mediating regulatable efflux because it is regulated by InsPPs through its SPX domain (Potapenko et al, 2018, 2019; Wang et al, 2015; Hürlimann et al, 2007). Furthermore, an overexpressed GFP fusion was reported to localize to vacuoles (Hürlimann et al, 2007; Elbaz-Alon et al, 2014), a phenotype that we confirmed (not shown). Since overexpression of membrane proteins in yeast easily leads to their erroneous accumulation in vacuoles, we re-investigated Pho91 localization in the absence of a strong, overexpressing promotor. When we expressed C- or N-terminal Pho91-GFP fusions from the authentic PHO91 promotor, we observed different and complex localization patterns (see Supplementary Figure 2 for examples).

These localization patterns depended on the nature of the fluorescent tag and the linker peptides used to attach it to Pho91, and they often showed significant accumulation in the ER. This suggests that the fluorescent protein tags interfere with intracellular trafficking of Pho91. We hence analysed purified vacuoles to test whether the non-tagged, endogenous Pho91 is a vacuolar protein. Mass spectrometry was used to determine the enrichment of proteins in the vacuolar fraction relative to a whole cell extract (Suppl. Table 1). Peptides from Pho91 were enriched in the purified vacuoles to a similar degree (39-fold) as peptides from vacuolar marker proteins, such as the vacuolar polyP polymerase subunit Vtc3 (35-fold), the vacuolar amino acid transporter Avt3 (48-fold), the v-ATPase subunit Vma9 (46-fold), or the alkaline phosphatase Pho8 (34-fold). By contrast, typical plasma membrane proteins were barely enriched, such as the iron permease Ftr1 (2.5-fold), the Pi importer Pho87 (2.4-fold) or the polyamine importer Tpo5 (2.2-fold) (Supplementary data file 1). This co-enrichment suggests that the major fraction of non-tagged Pho91 indeed resides in the vacuole. In agreement with this, Pho91 engages the AP3-dependent vesicular trafficking pathway, which leads from the Golgi to the vacuole (Eising et al, 2022). In contrast to the other four yeast Pi transporters, Pho84, Pho87, Pho89, and Pho90, which individually suffice to support vigorous growth of yeast, Pho91 was reported to do support slower but significant growth and to fully rescue upon overexpression (Wykoff & O’Shea, 2001; Wykoff et al, 2007). Since this capacity to feed the cells is difficult to reconcile with a vacuolar localisation, unless the cells would feed through fluid phase endocytosis and subsequent export from the vacuole (Klompmaker et al, 2017), we reinvestigated this aspect. Instead of relying a down-regulatable PHO84 background, which results in a low but constitutive background expression of the potent Pi importer Pho84 in addition to Pho91, we relied on a more stringent approach, using plasmid shuffling to completely remove PHO89 when PHO91 is brought in. To this end, we generated a quintuple knockout strain lacking these five Pi transporters (Pho84, Pho87, Pho89, Pho90, Pho91), which was kept alive through expression of Pho89 from a URA-based plasmid. Plasmid shuffling allowed to exchange this plasmid against others expressing an individual Pi transporter. Whereas an individual plasma-membrane Pi transporter such as Pho89 supported normal colony formation, confirming previous observations (Wykoff et al, 2007; Wykoff & O’Shea, 2001), we could not confirm that Pho91 supports slower growth. In our plasmid shuffling approach, Pho91 did not support growth, at all (Suppl. Fig. 3). This supports that Pho91 is indeed a vacuolar Pi transporter, but that C- or N-terminal protein fusions interfere with correct sorting and cannot serve as reliable reporters for localization of this protein.

We tested the impact of Pho91 on Pi accumulation by vacuoles in vitro, using the same approach as above (Fig. 4). Vacuoles lacking Pho91 (pho91Δ) accumulated Pi two times faster than the wildtype and the maximal accumulated concentration was two times higher. This is consistent with a function of Pho91 as vacuolar Pi exporter (Hürlimann et al, 2007; Potapenko et al, 2018; Wang et al, 2015).

Accumulation of Pi in isolated vacuoles

VTC- and Pho91-dependence. Vacuoles were isolated from the indicated, logarithmically growing strains. The purified organelles were incubated as in Fig. 1, i.e. in a buffer with an ATP-regenerating system that allows the synthesis of polyP, and either (A) without (w/o) or (B) in the presence of 50 µM 5-IP7. After the indicated periods of incubation at 27°C, an 80 µl aliquot was withdrawn, the vacuoles were sedimented by centrifugation, washed and lysed. Released vacuolar Pi was determined by malachite green assay. Graphs represent the mean ± SEM of three independent experiments. p<0.05 from a paired t-test comparing each time WT versus pho91Δ.

The difference between pho91Δ and wildtype vacuoles vanished when the vacuoles were incubated in the presence of the inositol pyrophosphate 5-InsP7. Since, as we showed above, vacuolar Pi accumulation depends on polyP synthesis through VTC, the relative enhancement of Pi accumulation in wildtype vacuoles through 5-InsP7 could be caused by downregulation of Pho91, which would make the wildtype vacuoles behave similarly as pho91Δ vacuoles. Alternatively, 5-InsP7 might stimulate polyP synthesis in wildtype more than in pho91Δ vacuoles. We could rule out the latter explanation based on two observations: An assay of polyP synthesis during the incubation (Fig. 5) revealed that 5-InsP7 stimulated polyP synthesis in wildtype and pho91Δ vacuoles to similar degrees. Furthermore, as shown above (Fig. 2), polyP synthesis activity in wildtype cells is not rate-limiting for vacuolar Pi accumulation. Therefore, we attribute the enhancement of Pi accumulation through 5-InsP7 to Pho91. Upon Pi scarcity, the declining inositol pyrophosphate levels should then activate Pho91 to replenish cytosolic Pi.

PolyP accumulation in pho91 mutant vacuoles.

Vacuoles from BY4742 cells (WT) and isogenic pho91Δ cells were isolated and incubated under conditions supporting polyP synthesis and Pi accumulation as in Fig. 4, in the absence (w/o) or presence of 50 µM 5-InsP7. At the indicated timepoints, aliquots were withdrawn, the vacuoles were lysed in detergent, and polyP was assayed through DAPI fluorescence. Graphs show the means and SEM from three independent experiments. p>0.05 from a paired t-test for all differences between WT and pho91Δ.

Interference with vacuolar polyP turnover provokes cytosolic Pi scarcity

We tested the effects of this postulated cycle of polyP synthesis, polyphosphatase activity and Pho91-mediated Pi export on Pi homeostasis in vivo. Pho4-GFP was used as a reporter, because this transcription factor shuttles between nucleus and cytosol. Under cellular Pi scarcity and correspondingly low inositol pyrophosphate levels it is predominantly nuclear, but it shifts to the cytosol under Pi-replete conditions (O’Neill et al, 1996; Chabert et al, 2023; Desfougères et al, 2016a; Auesukaree et al, 2004). Pho4-GFP localisation can hence serve as a readout for cytosolic Pi signalling. In wildtype cultures growing logarithmically, cells can experience a transient shortage of Pi during S-phase, when Pi utilization for biosynthesis may exceed the uptake capacity of the cell and consume its polyP stores (Bru et al, 2016; Neef & Kladde, 2003; Pondugula et al, 2009). Therefore, even on Pi-rich media, some cells can show nuclear accumulation of Pho4-GFP (Wykoff et al, 2007; Vardi et al, 2013, 2014). Since S-phase occupies only around 20% of the yeast cycle (Koren et al, 2010; Donaldson et al, 1998), this fraction is expected to be correspondingly small. However, it is significant, and it can be clearly quantified by microscopy, which scores individual cells and not the population as an ensemble. We exploited this natural fluctuation of intracellular Pi availability as a sensitising situation to assay how the vacuolar systems influence Pi balance in the cytosol.

We scored the fraction of logarithmically growing cells that show nuclear Pho4-GFP even in Pi-replete medium. This fraction was 7% for wildtype cells (Fig. 6). It increased fourfold upon hyperactivation of polyP synthesis, which we achieved by overexpression of the regulatory VTC subunit Vtc5 (Desfougères et al, 2016a). Ablation of polyP synthesis by deletion of the catalytic subunit Vtc4 had the opposite effect and reduced the frequency of cells with nuclear Pho4-GFP by half. Deletion of the vacuolar polyphosphatases Ppn1 and Ppn2 prevents polyP turnover and leads to the accumulation of extremely long polyP chains (Gerasimaite & Mayer, 2017; Sethuraman et al, 2001). We may thus expect phosphate to remain fixed in the form of polyP instead of being made available for Pi reflux into the cytosol.

Impact of the vacuolar polyP/Pi cycle on cytosolic Pi signalling

A) Illustration of nucleo-cytoplasmic relocation of Pho4-GFP in response to Pi availability. Wildtype yeast cells were grown in SC medium under Pi replete conditions. During exponential phase (OD600nm=1), cells were transferred for 30 min to synthetic complete media with 200 µM phosphate (-Pi) or 7.5 mM Pi (+Pi) and imaged by fluorescence microscopy. B) The indicated yeast strains were logarithmically grown over night in SC medium with 7.5 mM Pi,, harvested at OD600nm=1 and immediately imaged by fluorescence microscopy as in A. The graph shows the means and SEM of the percentage of cells showing Pho4 predominantly in the nucleus. n=3 independent experiments with 200 cells quantified per sample. ** p<0.005, *** p<0.001, ****p<0.0001 from a paired t-test comparing WT with the mutants.

In line with this, the fraction of ppn1Δ ppn2Δ double mutants that showed nuclear Pho4-GFP was 3-fold higher than in wildtype. pho91Δ cells, in which we expect Pi export from the vacuoles to be impaired, showed two times higher frequency of nuclear Pho4-GFP than wildtype. pho91Δ vtc4Δ double mutants showed an even lower frequency of nuclear Pho4-GFP than wildtype, suggesting that the state of Pi starvation that pho91Δ favours is dependent on vacuolar polyP accumulation. Cells expressing pho91K237A as the sole source of Pho91 also showed a 50% lower frequency of nuclear Pho4-GFP. The pho91K237A allele generates an amino acid substitution in the InsPP-binding patch of the Pho91 SPX domain. It mimics the InsPP-free state (Wild et al, 2016) and hence low-Pi conditions (Chabert et al, 2023). Collectively, our results are consistent with the notion that loss of InsPP binding activates Pho91 to export Pi from vacuoles to the cytosol, enhancing repression of the PHO pathway.

Since InsPP levels in yeast decline in response to cellular Pi availability (Lonetti et al, 2011; Chabert et al, 2023), we tested the impact of Pho91, Ppn1 and Ppn2 on the cellular levels of these metabolites using CE-MS, capillary electrophoresis coupled to mass spectrometry (CE-MS) (Qiu et al, 2020, 2023, 2021). The cells were harvested from the same Pi-replete growth conditions as for the microscopic assays above (Fig. 7).

Impact of Pho91 and vacuolar polyphosphatases on InsPP levels.

pho91Δ and ppn1Δppn2Δ cells, as well as their isogenic wildtypes, were logarithmically grown in Pi-replete SC medium as in Fig. 6. At OD600nm=1 the cells were extracted with perchloric acid as previously described (Wilson et al, 2015) and analysed for the indicated InsPs through capillary electrophoresis coupled to mass spectrometry (CE-MS) as described (Qiu et al, 2023). n=3. * p<0.05 from a paired t-test comparing WT with the mutants.

The ppn1Δ ppn2Δ double mutants showed a decrease by 30 to 50% in the inositol pyrophosphates 1,5-InsP8, 5-InsP7 and 1-InsP7, whereas InsP6 remained at a similar level as in wildtype. InsPP changes at this scale are functionally significant, because they suffice to trigger the initial phase of the Pi starvation response (Chabert et al, 2023; Kim et al, 2023, 2025). This is consistent with the 3-fold increase of cells with nuclear Pho4-GFP (Fig. 6). pho91Δ cells did not show significant changes for all four metabolites although they showed a partial shift of Pho4-GFP into the nucleus (Fig. 6). We attribute this discrepancy to the different nature of the assays. The microscopic assay for Pho4-GFP localization can pick up effects in a fraction of the cells because it offers single cell resolution. Inositol pyrophosphate analysis is an ensemble assay, in which changes in a smaller fraction of the population become diluted through the major pool that does not show the effect. For this reason, a change in inositol pyrophosphates affecting only 10% of the pho91Δ cells that show nuclear Pho4-GFP remains undetectable. The microscopic assay suggests that cytosolic Pi scarcity in ppn1Δ ppn2Δ affects more cells, perhaps because it is more profound, and it hence becomes detectable even in a whole-population-analysis.

Discussion

Our observations can be integrated with existing data on the properties of VTC (Wild et al, 2016; Hothorn et al, 2009; Müller et al, 2002; Gerasimaite et al, 2014, 2017; Liu et al, 2023; Guan et al, 2023; Pipercevic et al, 2023) and Pho91 (Hürlimann et al, 2007; Wang et al, 2015; Potapenko et al, 2018) to generate a working model explaining how an acidocalcisome-like organelle such as the yeast vacuole is set up to function as a Pi buffer for the cytosol. Under Pi-replete conditions, high InsPP levels activate VTC to polymerize Pi into polyP and translocate it into the vacuolar lumen. Here, the vacuolar polyphosphatases degrade polyP into Pi, filling the lumen with Pi (Sethuraman et al, 2001; Shi & Kornberg, 2005; Lichko et al, 2010; Gerasimaite & Mayer, 2017). Since the Pi exporter Pho91 is downregulated through InsPP binding to its SPX domain (Potapenko et al, 2018; Hürlimann et al, 2007; Wang et al, 2015), the Pi liberated through polyP hydrolysis accumulates in the vacuoles. Product inhibition of the polyphosphatases attenuates polyP hydrolysis once the vacuolar lumen has reached a Pi concentration above 30 mM. When the cells experience Pi scarcity, InsPP levels decline (Chabert et al, 2023). This activates Pho91 to release Pi from the vacuolar pool into the cytosol and stabilizes cytosolic Pi.

Working model of acidocalcisome-like vacuoles as Pi buffering systems.

(A) Under Pi-replete conditions, ATP drives the conversion of Pi into polyP and its translocation into the organelle. Here, polyP is degraded by the vacuolar polyphosphatases Ppn1 and Ppn2 to establish a vacuolar pool of free Pi. Feedback inhibition of Pi gradually reduces polyP degradation, enabling the buildup of a vacuolar polyP stock. Red lines and SPX colouring indicate inhibitory action, green colouring stimulation. (B) Cytosolic Pi scarcity decreases InsPP levels, which triggers two compensatory, SPX-controlled effects: The transfer of Pi from the cytosol into vacuoles through VTC ceases; and Pho91-dependent export of Pi from vacuoles is activated. Both measures synergize to stabilize cytosolic Pi. The export of Pi from the vacuole lifts product inhibition on the polyphosphatases Ppn1 and Ppn2 and stimulates a compensatory degradation of polyP.

Yeast cells do not only accumulate Pi as a rapidly accessible buffer for the cytosol. Under Pi-replete conditions they accumulate hundreds of millimolar of phosphate in the form of polyP (Urech et al, 1978). In contrast to the vacuolar reserve of Pi, which is presumed to be accessible immediately, mobilizing the polyP store takes minutes to hours (Nicolay et al, 1982; Pondugula et al, 2009; Bru et al, 2016). But the polyP store offers advantages in the form of high capacity - hundreds of millimolar of phosphate units can be stored in the form of polyP - and low osmotic activity of polyP (Dürr et al, 1979). Keeping such a large stock of a critical resource, which is often growth-limiting in nature, is relevant for the cells. In case of phosphate shortage, the vacuolar polyP store can be mobilized to enable the cells to complete the cell cycle and transition into G0 phase (Spain DCB Facultat de Medicina i Ciencies de la Salut Universitat Internacional de Catalunya, 2015; Westenberg et al, 1989; Müller et al, 1992). This can consume substantial amounts of phosphate, because we can estimate that replicating the entire DNA (1.2*107 base pairs) immobilizes roughly 1 mM Pi in the cells; and cellular RNA is even 50 times more abundant than DNA (Warner, 1999), accounting for 50 mM phosphate. Phospholipids, which must also be synthesized to complete a cell cycle, fix phosphate in similar amounts (Lange & Heijnen, 2001). Thus, a large polyP store is necessary to guarantee that the cells can finish S-phase upon a shortage of phosphate sources. In accord with this notion, absence of the polyP store impairs cell cycle progression, nucleotide synthesis and induces genome instability (Bru et al, 2016, 2017). Also, a shift from non-fermentable carbon sources to fermentation of glucose leads to a strong requirement for Pi because the activation of glucose uptake and glycolysis depends on large amounts of phosphate-containing sugars and glycolytic intermediates (Nicolay et al, 1982, 1983; Gillies et al, 1981). Shortage of Pi restrains the abundance of these metabolites (Kim et al, 2023).

The properties of the regulatory circuit described above imply an inbuilt switch from vacuolar Pi accumulation to large-scale stocking of vacuolar polyP. Pi-replete conditions generate high cellular InsPP levels. These will not only reduce Pi efflux from the vacuoles through Pho91 and inactivate the vacuolar polyphosphatases, but at the same time stimulate continued polyP synthesis by VTC. Coincidence of these effects will favour storage and high accumulation of phosphate in the form of polyP. Conversely, depletion of the vacuolar Pi reservoir upon Pi scarcity in the medium will activate the vacuolar polyphosphatases. In combination with the downregulation of the polyP polymerase VTC through the decline of InsPPs this will mobilize the large vacuolar polyP reserve once the immediately available vacuolar Pi pool is gradually depleted.

The concentration of Pi inside vacuoles as a rapidly accessible Pi reserve, and the synthesis of a large polyP stock, come at an energetic cost because the transformation of Pi into polyP requires the formation of phosphoric anhydride bonds (Hothorn et al, 2009; Gerasimaite et al, 2014) and vacuolar Pi reaches 30 mM. This exceeds the cytosolic Pi concentration, which was measured through 31P-NMR in a variety of yeasts, yielding values of 5-17 mM (Nicolay et al, 1982, 1983). Cytosolic Pi can also be estimated based on data from several other studies (Auesukaree et al, 2004; Hürlimann et al, 2009; Theobald et al, 1996; Zhang et al, 2015; Pinson et al, 2004). Assuming that the cytosolic volume of a BY4741 yeast cell is 40 fL (Chabert et al, 2023), and 1 g of dry weight contains 40 * 109 yeast cells, these studies point to cytosolic values of 10-15 mM in Pi-replete media. Upon Pi starvation, this value rapidly drops up to fivefold, resulting in a strong Pi gradient across the vacuolar membrane (Okorokov et al, 1980; Shirahama et al, 1996). To replenish the cytosolic pool under Pi scarcity, Pho91 can exploit not only this Pi concentration gradient, but also the vacuolar electrochemical potential, which was shown to stimulate Pi export through the Pho91 homolog OsSPX-MFS3 from plant vacuoles (Wang et al, 2015).

Vacuolar Pi accumulation is driven indirectly through ATP in two ways. VTC uses ATP as a substrate and transfers the phosphoric anhydride bond of the ψ-phosphate onto a polyP chain (Hothorn et al, 2009). The growing polyP chain exits from the catalytic site directly towards the transmembrane part of VTC (Liu et al, 2023; Guan et al, 2023). This transmembrane part likely forms a controlled channel that can guide polyP through the membrane (Liu et al, 2023). Coupled synthesis and translocation require the V-ATPase (Gerasimaite et al, 2014), probably because polyP is highly negatively charged and therefore follows the electrochemical potential across the vacuolar membrane of 180 mV (inside positive) and 1.7 pH units (Kakinuma et al, 1981), which is generated through the proton pumping V-ATPase. Thus, the combination of the VTC complex and vacuolar polyphosphatases can be considered as a Pi pump that is driven by ATP through polyP synthesis and through the electrochemical potential for polyP translocation and Pi export.

It is likely that acidocalcisome- and lysosome-like organelles of other organisms act as buffers for cytosolic Pi similarly as described in our model for yeasts. This notion is supported by the conserved molecular setup of acidocalcisome-like organelles as well as by phenotypic similarities. The acidocalcisomes of Trypanosomes contain VTC, a Pho91 homolog and proton pumps in their membranes, and polyphosphatases in their lumen (Huang & Docampo, 2015; Billington et al, 2023; Fang et al, 2007; Lander et al, 2013; Ulrich et al, 2013; Scott et al, 1997). Also the acidocalcisome-like organelles of the alga Chlamydomonas contain such proteins and they accumulate polyP through VTC as a function of the availability of Pi, a proton gradient and metal ions (Zúñiga-Burgos et al, 2024; Blaby-Haas & Merchant, 2014; Long et al, 2023; Hong-Hermesdorf et al, 2014; Aksoy et al, 2014; Ruiz et al, 2001; Goodenough et al, 2019). Like in yeast, the polyP stores are mobilized upon Pi limitation (Sanz-Luque et al, 2020; Plouviez et al, 2021). Drosophila has a potentially lysosome-related compartment, which is acidic, carries V-ATPase and a homolog of the Pi exporter XPR1, impacts cytosolic Pi and diminishes upon Pi starvation (Xu et al, 2023). Mammalian lysosome-like organelles also participate in Pi homeostasis. They can accumulate polyP and take up Pi (Pisoni, 1991; Pisoni & Lindley, 1992). They carry the Pi exporter XPR1, which interacts with the plasma membrane Pi importer PiT1 to regulate its degradation (Li et al, 2024).

We hence propose that acidocalcisome-like vacuoles may have a general role as feedback-controlled, rapidly accessible Pi buffers for the cytosol, addressing a critical parameter for metabolism. However, given that acidocalcisome-like organelles accumulate not only phosphate but also multiple other metabolites and ions (Docampo, 2024), they are probably interlinked with cellular metabolism in multiple ways and might form an important hub for its homeostasis.

Materials and Methods

Yeast strains and growth conditions

Saccharomyces cerevisiae cells were grown on yeast extract-peptone-dextrose (YPD: 1% yeast extract, 2% peptone and 2% dextrose) or in synthetic complete (SC) medium from Formedium, supplemented with sodium phosphate as needed). Yeast backgrounds used in this study were BY4741 and BY4742. Genetic manipulations of yeast were performed by homologous recombination according to published procedures and/or transformation with the indicated plasmids (Gietz & Schiestl, 2007; Güldener et al, 1996). Strains used are listed in Table 1.

PHO4 localization

Cells transformed with pRS415-PPHO4-PHO4-GFP, a plasmid expressing a PHO4-GFP fusion from the PHO4 promotor (Chabert et al, 2023), were grown exponentially overnight in SC medium without leucine (SC-Leu). Care was taken that the culture did not grow beyond a density of OD600=0.7. At this point, the cells were collected by brief centrifugation (15 sec, 3000 x g) in a tabletop centrifuge and resuspended in 1/10th to 1/20th of their own supernatant. Pho4-GFP localization was immediately checked by microscopy. An aliquot of the cells was washed twice with SC-Leu without Pi and then diluted in SC-Leu with 200 μM phosphate to OD600=0.7. After 30 minutes of incubation in this medium, Pho4-GFP localization was analysed by fluorescence microscopy on a LEICA DMI6000B inverted microscope equipped with a Hamamatsu ORCA-R2 (C10600-10B) camera, an XCite ® series 120Q UV lamp and a Leica 100x 1.4 NA lens.

Vacuole preparation

Vacuoles were purified from yeast cells essentially as described (D’Agostino & Mayer, 2019). Briefly, yeast cells were grown in 1L of YPD to an OD600 of 1.5. 330 ml of cells were collected by centrifugation and resuspended in 50 ml of 30 mM Tris-HCl pH 8.9, 10 mM DTT buffer. Suspensions were incubated in a 30° C water bath for 5 min and collected by centrifugation. The pellet was resuspended in 15 ml of spheroplasting solution (50 mM K-phosphate pH 7.5, 600 mM sorbitol in YPD with 0.2% D-glucose and 3600 U/ml lyticase) and incubated for 25 min at 30°C. Spheroplasts were collected by centrifugation (2500 x g, 3 min) and resuspended in 2.5 ml of 15% Ficoll 400 in PS buffer (10 mM PIPES-KOH pH 6.8, 200 mM sorbitol). 80 μg of DEAE-dextran were added under gentle mixing. After incubation on ice for 2 min and then at 30 °C for 80 sec, spheroplasts were transferred into Beckman SW41.1 tubes, overlaid with cushions of 8%, 4% and 0% Ficoll 400 in PS buffer, and centrifuged (150’000 x g, 90 min, 2 °C). Vacuoles were collected from the 0-4 % Ficoll interface. Their protein concentration was determined through Bradford assay using BSA as a standard.

Polyphosphatase activity of vacuolar lysates

Polyphosphatase activity was assayed as described previously (Gerasimaite & Mayer, 2017), with the following modifications. Isolated vacuoles were diluted to a final protein concentration of 0.002 mg/ml in 1 ml of reaction buffer (20 mM PIPES/KOH pH 6.8, 150 mM KCl, 1 mM ZnCl2 or MgCwwl2, 0.1% Triton X-100, 1xPIC, 1 mM PMSF and 30 or 300 µM polyP300) in the presence or absence of various concentrations of KH2PO4 and incubated at 27°C. At the indicated times, 80 µl aliquots were collected and the reaction was stopped by dilution with 160 µl of stop solution (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 12 mM EDTA pH 8.0, 0.1% Triton X-100, 15 µM DAPI). Remaining polyP was quantified by measuring polyP-DAPI fluorescence (λexc. 415 nm, λem. 550 nm) in a black 96 well plate in a Spectramax Gemini microplate fluorometer (Molecular Devices). A reaction containing boiled vacuoles was used as negative control.

Polyphosphate synthesis by isolated vacuoles

Polyphosphate synthesis was assayed as described (Gerasimaite et al, 2014). Isolated vacuoles were diluted to final protein concentration of 0.02 mg/ml on 1 ml of reaction buffer (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 0.5 mM MnCl2, 200 mM sorbitol) and the reaction was started by adding an ATP regenerating system (1 mM ATP-MgCl2, 40 mM creatine phosphate and 0.25 mg/ml creatine kinase). The mix was incubated at 27 °C. At different timepoints, 80 µl aliquots were mixed with 160 µl of stop solution (10 mM PIPES/KOH pH 6.8, 150 mM KCl, 200 mM sorbitol, 12 mM EDTA, 0.15% Triton X-100 and 15 μM DAPI). PolyP synthesis was quantified through polyP-DAPI fluorescence (λexc. 415 nm, λem. 550 nm) in a black 96 well plate. A calibration curve was prepared using commercial polyP60 as a standard.

Phosphate quantification in isolated vacuoles

Isolated vacuoles were incubated as described for the polyP synthesis assay above. At different time points, 80 µl aliquots were centrifuged (3 min, 2000 x g, 2°C), the pellets were washed with 500 µl of washing solution (10 mM PIPES/KOH pH 6.8, 200 mM Sorbitol, 150 mM KCl) and centrifuged as before. The final pellet was resuspended with 100 µl of lysis buffer (10 mM PIPES/KOH pH 6.8, 200 mM sorbitol, 150 mM KCl, 12 mM EDTA, 0.1 % Triton). Free phosphate was quantified by adding 150 µl of molybdate-malachite green solution (1 mM malachite green, 10 mM ammonium molybdate,1M HCl) and reading the absorbance at 595 nm in a microplate photometer.

Inositol pyrophosphate synthesis, extraction and quantification

5-InsP7 was synthesized as described (Capolicchio et al, 2013; Wang et al, 2014). For quantification from cells, InsPPs extraction was performed as described previously (Kim et al, 2023), with the following modifications. Briefly, 3 ml of yeast cell culture at OD600nm=1 was collected using rapid vacuum-filtration on a polytetrafluoroethylene membrane filter (1.2 μm; Piper Filter GmbH, Germany). After snap freezing in liquid nitrogen, yeast cells on the membrane were resuspended in 400 μL of 1 M perchloric acid and lysed by bead beating (glass beads; 0.25-0.5 mm) for 10 min at 4 °C. After centrifugation at 13,000 rpm for 3 min at 4 °C, the supernatant was transferred into a new tube containing 3 mg of titanium dioxide (TiO2) beads (GL Sciences, Japan) which had been pre-washed twice with H2O and 1 M perchloric acid. The sample was gently rotated for 15 min at 4 °C. The TiO2 beads were collected by centrifugation at 13,000 rpm for 1 min at 4 °C and washed twice using 1 M perchloric acid. After the second washing step, the TiO2 beads were resuspended in 300 μL of 3 % (v/v) NH4OH and rotated gently at room temperature. After centrifugation at 13,000 rpm for 1 min, the eluants were transferred into a new tube and dried in SpeedVac (Labogene, Denmark) at 42 °C. InsPPs were measured through capillary electrophoresis coupled to mass spectrometry (CE-MS) as described (Qiu et al, 2021).

Quantification of the vacuolar proteome

Vacuoles were prepared as described above. During the purification procedure, a sample of the sedimented spheroplasts was withdrawn before DEAE dextran was added. These withdrawn spheroplasts constitute the “whole cell” extract. After the flotation step, the vacuoles withdrawn from the 4%-0% Ficoll interface were used as the vacuole fraction. For each fraction samples with 100 µg protein were precipitated by adding a final concentration 12.5% TCA for 10 min on ice. The proteins were sedimented (12’000 x g, 5 min, room temperature), the supernatant discarded, and the pellets were washed with ice-cold acetone and centrifuged as above twice. The final pellet was dried and dissolved in reducing SDS sample buffer (10% glycerol, 50 mM Tris-HCl pH 6.8, 2 mM DTT, 2% SDS, 0.002% bromophenol blue).

Samples dissolved in Laemmli buffer (50 mM Tris pH 6.8, 10 mM DTT, 2 % SDS, 0.1 % bromophenol blue, 10 % glycerol) were equalized in concentration by dilution in SP3 buffer (2% SDS, 10 mM DTT, 50 mM Tris, pH 7.5). 150 µg of protein per sample were heated at 95 °C for 5 min and cooled down. Reduced cysteine residues were alkylated by adding iodoacetamide (30 mM final) and incubating for 45 min at room temperature in the dark. Digestion was done by the SP3 method (Hughes et al, 2019) using magnetic Sera-Mag Speedbeads (Cytiva 45152105050250, 50 mg/ml). Beads were added at a ratio 10:1 (w:w) to samples, and proteins were precipitated on beads with ethanol (final concentration: 60 %). After 3 washes with 80% ethanol, beads were digested in 50 μl of 100 mM ammonium bicarbonate with 3.0 μg of trypsin (Promega #V5073). After 1 h of incubation at 37 °C, the same amount of trypsin was added to the samples for an additional 1 h of incubation. Supernatants were then recovered, transferred to new tubes, acidified with formic acid (0.5% final concentration), and dried by centrifugal evaporation. To remove traces of SDS, two sample volumes of isopropanol containing 1% TFA were added to the digests, and the samples were desalted on a strong cation exchange (SCX) plate (Oasis MCX; Waters Corp., Milford, MA) by centrifugation. After washing with isopropanol/1%TFA and 2% acetonitrile/0.1% FA, peptides were eluted in 200µl of 80% MeCN, 19% water, 1% (v/v) ammonia, and dried by centrifugal evaporation. Data-dependent LC-MS/MS analyses of samples were carried out on a Fusion Tribrid Orbitrap mass spectrometer (Thermo Fisher Scientific) interfaced through a nano-electrospray ion source to an Ultimate 3000 RSLCnano HPLC system (Dionex). Peptides were separated on a reversed-phase custom packed 45 cm C18 column (75 μm ID, 100Å, Reprosil Pur 1.9 μm particles, Dr. Maisch, Germany) with a 4-90% acetonitrile gradient in 0.1% formic acid at a flow rate of 250 nl/min (total time 140 min). Full MS survey scans were performed at 120’000 resolution. A data-dependent acquisition method controlled by Xcalibur software (Thermo Fisher Scientific) was used that optimized the number of precursors selected (“top speed”) of charge 2+ to 5+ while maintaining a fixed scan cycle of 0.6 s. Peptides were fragmented by higher energy collision dissociation (HCD) with a normalized energy of 32%. The precursor isolation window used was 1.6 Th, and the MS2 scans were done in the ion trap. The m/z of fragmented precursors was then dynamically excluded from selection during 60 s.

Data files were analysed with MaxQuant 1.6.14.0 (Cox & Mann, 2008) incorporating the Andromeda search engine (Cox et al, 2011). Cysteine carbamidomethylation was selected as fixed modification while methionine oxidation and protein N-terminal acetylation were specified as variable modifications. The sequence databases used for searching were the S. cerevisiae reference proteome based on the UniProt database (https://www.uniprot.org, version of June 6th, 2021, containing 6050 sequences), and a “contaminant” database containing the most usual environmental contaminants and enzymes used for digestion (keratins, trypsin, etc). Mass tolerance was 4.5 ppm on precursors (after recalibration) and 20 ppm on MS/MS fragments. Both peptide and protein identifications were filtered at 1% FDR relative to hits against a decoy database built by reversing protein sequences. The match between runs feature was enabled.